---
title: |
  Linkage Mapping and QTL Analysis for Plant Height in a Maize
  Recombinant Inbred Line Population Derived from the Nested
  Association Mapping Panel
author: |
  Brown Yomi Wilson  
  Graduate Research Intern, Germplasm Health/Virology and Molecular Diagnostic Unit, International Institute of Tropical
  Agriculture (IITA), Ibadan, Nigeria  
  Course: FIT678 -- Genetic Data Analysis in Plant Breeding  
  Federal University of Vicosa (UFV), Brazil
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    fig_caption: true
    fig_width: 9
    fig_height: 6
bibliography: references.bib
---

```{r global-options, include=FALSE}
# document-wide chunk settings, echo on so all code shows in the
# knitted report, warnings/messages off so package load chatter
# doesn't clutter the output, cache off so a full knit always
# reflects the current state of the saved objects
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  cache   = FALSE
)
```

---

# Abstract {-}

Plant height is an important agronomic trait in maize, influencing
lodging resistance, biomass accumulation, and overall crop
performance. This study aimed to construct a genetic linkage map
and identify quantitative trait loci (QTL) associated with plant
height in the Z020 recombinant inbred line (RIL) population, derived
from a cross between NC350 and B73 within the maize Nested
Association Mapping (NAM) panel. A linkage map spanning 1,548.7 cM
across 10 linkage groups was constructed from 242 SNP markers using
188 RILs genotyped with the MaizeSNP50 array. QTL mapping was
performed using a progression of four complementary methods: single
marker analysis, interval mapping, composite interval mapping, and
multiple interval mapping. Single marker analysis identified 13
markers significantly associated with plant height after Bonferroni
correction, with the strongest signal on linkage group 6 explaining
15.8% of phenotypic variance. Composite interval mapping confirmed
three robust QTL on linkage groups 4, 6, and 8, and multiple
interval mapping extended this to a final model of five QTL on
linkage groups 4, 5, 6, 8, and 10, jointly explaining 42.3% of
phenotypic variance. Four of the five QTL showed the NC350 allele
associated with increased height, while one locus on linkage group 5
showed the opposite, B73-favouring effect. No evidence of epistatic
interaction was detected among the identified loci, suggesting a
predominantly additive genetic architecture for plant height in this
population. These findings identify a small number of moderate to
large effect QTL suitable for further validation and potential use
in marker-assisted selection for plant height in maize breeding
programmes.

---

# Introduction

Plant height is one of the most extensively studied architectural
traits in maize, with direct consequences for crop performance.
Excessively tall plants are more prone to stalk and root lodging,
particularly under high planting density or adverse weather, while
plant height is also closely linked to biomass production, light
interception, and the partitioning of resources between vegetative
and reproductive growth. As a quantitative trait, plant height is
controlled by multiple genes of varying effect size, interacting
with environmental conditions, making it a useful model trait for
demonstrating the principles of linkage mapping and quantitative
trait loci (QTL) analysis in a crop species of major global
importance.

The maize Nested Association Mapping (NAM) panel was developed to
combine the advantages of linkage mapping and association mapping
in a single resource, by crossing a diverse set of inbred founder
lines to a common reference parent, B73, and advancing the progeny
through repeated selfing to produce recombinant inbred line (RIL)
populations. Each NAM family, such as the Z020 population used in
this study, derived from a cross between NC350 and B73, captures
the recombination history of a single biparental cross while
remaining part of a larger, structured panel that allows findings
to be compared and combined across many genetic backgrounds.
Recombinant inbred lines, having undergone several generations of
selfing, are particularly well suited to QTL mapping, since
individuals are expected to be nearly homozygous at every locus, which
simplifies the genetic model used in linkage and QTL analysis and
allows each line to be phenotyped repeatedly without concern for loss
of genetic fidelity between generations.

A genetic linkage map, built from the observed recombination
frequencies between molecular markers, provides the essential
scaffold for QTL analysis. By ordering markers along each chromosome
and converting recombination fractions into genetic distances, a
linkage map allows the genome to be scanned systematically for
regions associated with variation in a trait of interest. Once QTL
have been localised, the markers flanking them can be used directly
in marker-assisted selection, allowing breeders to select for
favourable alleles at a target locus without needing to phenotype
every individual in a breeding population, considerably accelerating
the breeding cycle for traits that are slow, costly, or destructive
to measure directly.

The objectives of this study were threefold: first, to construct a
genetic linkage map of the Z020 (NC350 x B73) RIL population using
SNP marker data from the MaizeSNP50 array; second, to identify QTL
associated with plant height using a progression of complementary
mapping approaches of increasing statistical sophistication, namely
single marker analysis, interval mapping, composite interval
mapping, and multiple interval mapping; and third, to characterise
the genetic architecture underlying plant height in this population,
including the number, effect size, and direction of effect of the
QTL identified, and to evaluate evidence for epistatic interaction
among them.

---

# Materials and Methods

## Plant Material and Mapping Population

The mapping population used in this study was the Z020 recombinant
inbred line (RIL) population, derived from a cross between NC350 and
B73, two inbred lines from the maize Nested Association Mapping (NAM)
panel [@McMullen2009]. The population consists of 188 RIL individuals
produced by repeated selfing from a single F2 plant, reaching
approximately the F6 to F7 generation. At this stage, residual
heterozygosity is expected to be approximately 5.5% of the genome.
The trait measured for QTL analysis was plant height (cm), recorded
from the base of the plant to the tip of the tassel at maturity.

## Molecular Markers and Genotypic Data

Genotypic data were generated using the MaizeSNP50 Illumina
BeadChip, which interrogates approximately 50,000 SNP loci across
the maize genome. After initial filtering, 1,106 polymorphic SNP
markers were retained for linkage analysis. Genotypes were coded as
0 for the NC350 homozygote (AA), 2 for the B73 homozygote (BB), and
1 for residual heterozygotes. Approximately 2,624 missing genotype
calls were present across the full marker set. For linkage map
construction, residual heterozygous calls (approximately 5.5% of
all genotype cells) were recoded as missing data, consistent with
the ri self cross type specification in the R/onemap package, which
does not accommodate a heterozygote class for near-fully inbred RIL
populations.

## Marker Segregation Analysis

Marker segregation was tested using the `test_segregation()` function
in R/onemap [@Taniguti2023]. For a self-derived RIL population, each
marker is expected to segregate at a 1:1 ratio of AA to BB
homozygotes. A chi-square goodness-of-fit test was applied to each
marker independently, and the experiment-wise significance threshold
was adjusted for multiple testing using the Bonferroni correction
(alpha / M, where M is the number of markers tested), yielding a
corrected threshold of approximately 0.0000452. Markers showing
significant segregation distortion were retained in the analysis,
as a moderate number of distorted markers is biologically expected
in RIL populations due to selection during selfing or allele-specific
fitness differences between the NC350 and B73 genetic backgrounds.

## Linkage Map Construction

Two-point recombination fractions were estimated between all possible
marker pairs using maximum likelihood via the `rf_2pts()` function in
R/onemap, with a LOD threshold of 5.916 and a maximum recombination
fraction of 0.35 used for marker grouping. Markers with identical or
near-identical segregation patterns across all individuals were
identified using `find_bins()` and reduced to a single representative
per bin prior to ordering, to prevent inflation of chromosome length
estimates. Marker ordering within each linkage group was performed
using the `order_seq()` function with a branch-and-bound algorithm
(n.init = 5, THRES = 3), and only markers placed with confidence
(the safe set) were retained in the final map. Multipoint genetic
distances were estimated using the Hidden Markov Model implemented
in `onemap::map()`, and recombination fractions were converted to
Kosambi centimorgans using the formula:
d = 0.25 x log[(1 + 2r) / (1 - 2r)] x 100.
The final linkage map and genotype data were exported for QTL
analysis using the R/qtl package [@Broman2003].

## QTL Mapping

### Single Marker Analysis

Single marker analysis (SMA) was performed by fitting a simple linear
regression model for each marker independently:

PlantHeight = mu + beta * x + epsilon

where mu is the population mean, x is the marker genotype coded as 0
(NC350 homozygote) or 1 (B73 homozygote), beta is the additive
effect of the B73 allele relative to NC350, and epsilon is the
residual error assumed to follow a normal distribution. LOD scores
were calculated from the F statistic of each regression as:
LOD = (N/2) x log10(F x (1/(N-2)) + 1).
The experiment-wise significance threshold was determined by
Bonferroni correction (alpha = 0.05 / M, where M is the number of
markers tested). Effect size at the most significant marker was
further estimated using the `fitqtl()` function in R/qtl with
multiple imputation (n.draws = 32). Results were cross-validated
using the `scanone(method = "mr")` function in R/qtl.

### Interval Mapping

Interval mapping (IM) was performed using the R/qtl package
[@Broman2003]. Conditional QTL genotype probabilities were calculated
at every 1 cM position along the linkage map using the
`calc.genoprob()` function, allowing the genome scan to evaluate
putative QTL positions between markers as well as at marker
locations. The genome scan was performed using both the
Expectation-Maximisation (EM) algorithm and Haley-Knott (HK)
regression via the `scanone()` function. Genome-wide significance
thresholds were determined empirically by 1,000 phenotype
permutations for each method separately, with significance declared
at alpha = 0.05. Effect size and the proportion of phenotypic
variance explained were estimated at the peak QTL position using
`fitqtl()` with multiple imputation (n.draws = 32).

### Composite Interval Mapping

Composite interval mapping (CIM) was performed using the `cim()` 
function in R/qtl, with Haley-Knott regression and a set of 27 marker 
cofactors selected automatically, following the convention of 
2 x square root(N) for a population of 188 individuals. Two window sizes 
were compared: 15 cM, in which cofactors within 15 cM of the position 
being tested were excluded from the model, and an infinite window, in 
which no cofactor from the same linkage group as the tested position 
was permitted regardless of distance. A fixed random seed (678) was used 
for both composite interval mapping runs to ensure reproducibility of marker 
cofactor selection. Genome-wide significance was 
assessed by 1,000 phenotype permutations using the infinite window 
setting, as this represents the most conservative cofactor exclusion 
rule and yields a threshold valid across both window sizes. Effect 
size and proportion of phenotypic variance explained were estimated 
using `fitqtl()` with multiple imputation (n.draws = 32), fitting all 
confirmed QTL jointly in a single additive model.

### Multiple Interval Mapping

Multiple interval mapping (MIM) was performed in R/qtl, building 
forward from the three QTL confirmed by composite interval mapping. 
QTL positions were jointly refined using `refineqtl()`, and additional 
QTL were sought iteratively using `addqtl()`, with each candidate 
position tested against a penalised LOD threshold of 2.55, derived 
from 1,000 two-dimensional genome-wide permutations (`scantwo()`) 
following the method of Manichaikul et al. (2009). After each 
addition, all QTL positions in the model were re-refined before the 
next round of scanning, and the search concluded once no further 
position exceeded the threshold. As an independent cross-check, an 
automatic stepwise model search was also performed using 
`stepwiseqtl()`, with a maximum of eight candidate QTL and the same 
permutation-derived penalties. Evidence for epistatic interaction was 
assessed in two ways: first using `addint()`, which tests pairwise 
interactions among QTL already included in the model, and second 
using a genome-wide two-dimensional scan (`scantwo()`), which can 
detect interacting loci regardless of their individual main effects. 
Interaction terms were evaluated against a penalised threshold of 
4.01. Effect sizes and proportion of phenotypic variance explained for 
the final model were estimated using `fitqtl()` with multiple 
imputation (n.draws = 32).

---

# Results

## Descriptive Statistics for Plant Height

Plant height in the Z020 RIL population ranged from approximately
89 cm to 202 cm across the 188 individuals, with a mean of
approximately 137 cm and a standard deviation of approximately 16 cm.
The distribution was approximately normal, as confirmed by a
Shapiro-Wilk test (W = 0.9893, p = 0.2956), indicating no
significant departure from normality at alpha = 0.05. This supports
the use of linear regression-based QTL mapping methods, which assume
normally distributed trait residuals. The approximately continuous
distribution of plant height values across the RIL population is
consistent with polygenic inheritance, where multiple loci of
moderate effect each contribute to the trait (Figure 1).

## Marker Segregation

Chi-square segregation tests were performed for all 1,106 SNP
markers under the expected 1:1 AA:BB ratio for a self-derived RIL
population. Using the Bonferroni-corrected significance threshold
of approximately 0.0000452, a subset of markers showed significant
deviation from the expected segregation ratio. These markers were
retained in the linkage analysis, as a moderate level of segregation
distortion is biologically expected in RIL populations derived from
crosses between genetically divergent inbred lines such as NC350 and
B73, and removal of distorted markers is not standard practice unless
distortion is extreme or clustered in a way that disrupts map
construction (Figure 3).

## Linkage Groups

All markers were assigned to linkage groups using a LOD threshold
of 5.916 and a maximum recombination fraction of 0.35. After
removal of redundant markers using `find_bins()`, the 1,106 SNP
markers were reduced to a set of unique segregation pattern
representatives. All assigned markers grouped into exactly 10
linkage groups, corresponding to the 10 chromosomes of the maize
genome (2n = 20). Linkage group sizes after bin filtering and
ordering ranged from 13 markers on linkage group 7 to 40 markers
on linkage group 9. The successful recovery of exactly 10 linkage
groups confirms that the marker density and population size of
the Z020 RIL population were sufficient for complete genome coverage
(Figure 4).

## Linkage Map

The final genetic linkage map comprised 242 SNP markers distributed
across 10 linkage groups, with a total map length of 1,548.7 cM
(Kosambi). Individual chromosome lengths ranged from 84.9 cM
(linkage group 5) to 256.7 cM (linkage group 9). The average
inter-marker spacing across the complete map was approximately
6.4 cM. The recombination fraction heatmap for linkage group 1
showed a clean diagonal band of low recombination fractions between
adjacent markers, confirming correct marker ordering (Figure 5).
Haplotype reconstruction for representative individuals revealed
the expected mosaic structure of large homozygous blocks alternating
between NC350 and B73 alleles, with colour transitions marking
historical recombination breakpoints accumulated during the selfing
generations (Figure 7). The complete 10-chromosome map is shown in
Figure 8, and a summary of marker counts and chromosome lengths is
presented in Table 1.

## QTL Analysis

### Single Marker Analysis

Single marker analysis identified 13 markers significantly
associated with plant height after Bonferroni correction
(threshold: LOD = 3.11, corresponding to P < 0.0000452 across
242 markers tested). The strongest association was detected at
marker PZA01142.4 on linkage group 6 (position 109.4 cM;
LOD = 6.62; R² = 0.158), indicating that this marker alone
explains 15.8% of the phenotypic variance in plant height.
At this marker, individuals homozygous for the NC350 allele
were on average 12.87 cm taller than those homozygous for the
B73 allele. A secondary peak was identified on linkage group 4
(LOD approximately 5.0), a borderline signal was present on
linkage group 5 (LOD approximately 3.0), and a notable peak
was detected on linkage group 10 at marker PZA02386.2
(LOD approximately 4.5). The direction of the allele effect
at linkage group 10 was opposite to all other significant
markers: at this locus, the B73 allele was associated with
taller plants (additive effect approximately +8.97 cm),
whereas the NC350 allele conferred the height advantage at
all other significant positions. No significant associations
were detected on linkage groups 1, 2, 3, 7, 8, or 9.
Effect size at the top marker was estimated using `fitqtl()`:
LOD = 6.87, variance explained = 15.49%, additive effect =
-6.357 cm (half-difference parameterisation, equivalent to
a full group difference of approximately -12.7 cm, consistent
with the regression estimate above). The LOD profile across
all 10 linkage groups is shown in Figure 10, and results for
the top marker are summarised in Table 2.

### Interval Mapping

Interval mapping identified four genomic regions significantly
associated with plant height at the genome-wide alpha = 0.05
level, based on permutation-derived LOD thresholds of 2.99
(EM method) and 2.79 (HK method) from 1,000 permutations
(Figure 11). The strongest signal was located on linkage
group 6 at 108 cM (LOD = 7.34, EM; LOD = 7.38, HK),
explaining 16.09% of the phenotypic variance in plant height.
The additive effect at this position was -6.49 cm
(half-difference parameterisation), corresponding to a full
group difference of approximately 13.0 cm, with NC350 alleles
associated with taller plants. Three additional QTL exceeded
the significance threshold: linkage group 4 at 67 cM
(LOD = 4.75, EM), linkage group 5 at 14.9 cM (LOD = 3.20,
EM), and linkage group 10 at 108 cM (LOD = 3.29, EM). The EM
and HK methods produced nearly identical LOD profiles and
identified the same set of significant positions, confirming
the robustness of the interval mapping results. Compared with
single marker analysis, interval mapping increased the peak
LOD score on linkage group 6 from 6.62 to 7.34 and the
estimated variance explained from 15.49% to 16.09%,
consistent with the expected improvement in resolution and
power when putative QTL positions between markers are
evaluated. A summary of all significant QTL positions
detected by interval mapping is presented in Table 2.


### Composite Interval Mapping

Composite interval mapping identified three genomic regions 
significantly associated with plant height, consistent across both 
window settings tested. Under the permutation-derived threshold of 
LOD 6.24 (alpha = 0.05, infinite window), significant QTL were 
detected on linkage group 4 at 66 cM (LOD = 14.58, window = infinite; 
LOD = 10.18, window = 15 cM), linkage group 6 at 108 cM 
(LOD = 18.58, window = infinite; LOD = 15.29, window = 15 cM), and 
linkage group 8 at 126.5 cM (LOD = 6.95, window = infinite; 
LOD = 7.57, window = 15 cM). The linkage group 6 signal remained the 
strongest detected at any stage of this analysis, with its LOD score 
increasing from 6.62 under single marker analysis to 7.34 under 
interval mapping and to 18.58 under composite interval mapping, 
indicating that controlling for background genetic variation 
substantially sharpened this signal. A QTL on linkage group 9, 
significant under interval mapping (LOD = 4.20), showed a 
window-dependent result under composite interval mapping: 
significant at the less conservative window = 15 cM setting 
(LOD = 7.38) but not at the more conservative infinite window 
(LOD = 4.33). This sensitivity to cofactor exclusion distance, in 
contrast with the stability of the linkage group 4, 6, and 8 signals 
across both window settings, suggests that part of the linkage 
group 9 signal may reflect incompletely controlled background 
variation rather than a fully independent genetic effect, and it 
was therefore not carried forward into the three-QTL model used for 
multiple interval mapping. The linkage group 8 QTL was not detected 
by either single marker analysis or interval mapping, and only 
emerged as significant once composite interval mapping accounted 
for the effects of the linkage group 4 and linkage group 6 loci, 
demonstrating the increased power of composite interval mapping to 
resolve QTL of moderate effect against a background of larger-effect 
loci elsewhere in the genome.

Fitting all three QTL jointly using `fitqtl()` explained 33.65% of 
phenotypic variance in plant height (LOD = 16.75). Each QTL 
contributed independently and significantly to the model, as shown 
by the drop-one-at-a-time analysis: the linkage group 4 QTL 
accounted for 13.21% of variance (LOD = 7.41, additive effect = 
-5.84 cm), the linkage group 6 QTL accounted for 18.85% of variance 
(LOD = 10.21, additive effect = -6.97 cm), and the linkage group 8 
QTL accounted for 4.42% of variance (LOD = 2.63, additive effect = 
-3.36 cm). At all three positions, the NC350 allele was associated 
with taller plants, consistent with the direction of effect observed 
at the major linkage group 6 QTL throughout single marker analysis 
and interval mapping. Individual contributions to variance explained 
do not sum to the full model total, since the drop-one-at-a-time 
percentages are each calculated relative to the complete three-QTL 
model and can therefore overlap. The complete CIM LOD profile across 
the genome, compared against interval mapping results for linkage 
groups 4, 6, and 8, is shown in Figure 13.

### Multiple Interval Mapping

Multiple interval mapping was initiated from the three QTL confirmed 
by composite interval mapping on linkage groups 4, 6, and 8. Joint 
refinement of these positions produced only a small shift relative to 
the composite interval mapping peaks (total LOD increase of 0.539), 
indicating that the composite interval mapping positions were already 
close to their jointly optimal locations. Iterative scanning for 
additional QTL using `addqtl()` identified two further loci exceeding 
the penalised threshold of 2.55: linkage group 10 at 102 cM 
(LOD = 2.99) and linkage group 5 at 13 cM (LOD = 2.65). No further QTL 
exceeded the threshold once these five positions were included in the 
model, and the manual search concluded with a final model of five QTL 
on linkage groups 4, 5, 6, 8, and 10.

The final five-QTL model explained 42.31% of phenotypic variance in 
plant height (LOD = 22.46). All five QTL remained individually 
significant in a drop-one-at-a-time analysis (Table 4). The linkage 
group 6 QTL (108 cM) contributed the largest individual effect 
(15.29% of variance, additive effect = -6.39 cm), followed by linkage 
group 4 (67 cM, 11.34% of variance, additive effect = -5.61 cm). 
Linkage groups 8 (126.5 cM) and 10 (102 cM) contributed more modestly 
(4.02% and 5.21% of variance respectively), and linkage group 5 
(13 cM) contributed the smallest individual effect (3.90% of 
variance). At linkage groups 4, 6, 8, and 10, the NC350 allele was 
associated with taller plants, consistent with the direction of 
effect observed throughout single marker analysis, interval mapping, 
and composite interval mapping. The linkage group 5 QTL showed the 
opposite direction of effect, with the B73 allele associated with 
greater plant height (additive effect = +3.26 cm), the only 
antagonistic locus identified in the final model.

An automatic stepwise model search (`stepwiseqtl()`), performed 
independently using the same permutation-derived penalties, selected 
a more conservative three-QTL model comprising linkage groups 4, 6, 
and 10 (pLOD = 9.59). This model omitted the linkage group 5 and 
linkage group 8 QTL identified by the manual search. The discrepancy 
reflects differences in search strategy rather than conflicting 
evidence: the manual search evaluated each candidate QTL against the 
addition of one locus to an existing model, while the stepwise search 
evaluated the overall penalised likelihood of the complete model 
globally. Linkage groups 4, 6, and 10 were identified as QTL by both 
approaches and are considered the most robustly supported loci in 
this analysis. Linkage groups 5 and 8 are reported as additional, 
more marginal QTL, supported by the manual search and by their 
individual significance in the five-QTL model, but not retained under 
the stricter global penalty applied by automatic stepwise selection. 
The LOD profile for the final five-QTL model is shown in Figure 14.

No evidence of epistatic interaction was found between QTL for plant 
height. Pairwise interaction tests among the five QTL in the final 
model (`addint()`) found a maximum interaction LOD of 0.36, well 
below the penalised threshold of 4.01. A genome-wide two-dimensional 
scan (`scantwo()`) similarly found no chromosome pair with an 
interaction LOD exceeding this threshold, the highest value observed 
being 3.00, between two positions on linkage group 7. These results 
indicate that the genetic architecture of plant height in this 
population is predominantly additive, with the five identified QTL 
acting independently rather than through interactive effects.

---

# Discussion

Five QTL were identified for plant height in the Z020 (NC350 x
B73) recombinant inbred line population, collectively explaining
42.31% of phenotypic variance. The two strongest loci, on linkage
groups 4 and 6, were detected consistently by every method applied
in this study, from single marker analysis through to multiple
interval mapping, with LOD scores and estimated effect sizes that
remained stable or strengthened as each successive method corrected
for background genetic variation. The linkage group 6 QTL in particular
proved to be the most robust finding of the entire analysis, with its LOD 
score rising from 6.62 under single marker analysis to 18.58 under composite 
interval mapping, indicating that, much of the noise obscuring this signal under
simpler methods was attributable to the combined effects of other loci elsewhere
in the genome rather than to weak linkage or low marker informativeness at
this locus.

The progression from single marker analysis to multiple interval
mapping also illustrates a more general point about QTL detection
that is easy to overlook when only the final model is reported.
The linkage group 8 QTL was undetectable by single marker analysis
and interval mapping, and only became visible once composite
interval mapping controlled for the variance explained by the
linkage group 4 and linkage group 6 loci. This is a direct
demonstration of the masking effect that larger-effect QTL can have
on smaller, independent loci elsewhere in the genome, and it
underlines why composite and multiple interval mapping are
considered more powerful tools than single marker or simple
interval mapping approaches for populations expected to carry more
than one causal locus.

A more cautious finding concerns the linkage groups 5 and 10 loci
and the direction of their effects. In the single marker analysis,
marker PZA02386.2 on linkage group 10 showed a positive effect of
the B73 allele on plant height, the only marker in that analysis to
contradict the otherwise consistent pattern of the NC350 allele
conferring greater height. Once a full multiple-QTL model was
fitted, however, the confirmed linkage group 10 QTL at 102 cM
showed the conventional, NC350-favouring direction of effect, while
the antagonistic, B73-favouring effect instead appeared at a newly
identified QTL on linkage group 5. Two explanations seem plausible.
It is possible that linkage groups 5 and 10 each carry genuinely
distinct QTL with opposing effects on plant height, and that single
marker analysis, lacking the ability to separate co-occurring
signals, attributed an effect to the wrong general region of the
genome. Alternatively, the single-marker estimate at linkage group
10 may simply have been a less reliable, more noise-prone estimate
that the better-controlled multiple-QTL model has now corrected.
Distinguishing between these explanations would require fine
mapping or independent validation in a related population, and is
beyond the scope of the present analysis, but the discrepancy is a
useful illustration of why effect direction, and not only effect
significance, should be interpreted cautiously when comparing
results across QTL mapping methods of differing sophistication.

The disagreement between the manually constructed five-QTL model
and the automatically selected three-QTL model from stepwiseqtl()
deserves comment rather than resolution in favour of one over the
other. Both searches used the same permutation-derived penalties
and the same underlying data, yet reached different conclusions
about whether the linkage group 5 and linkage group 8 loci belong
in the final model. This is not evidence that either method is
wrong, but rather a reflection of the different criteria each
applies: the manual search asks whether one additional locus
improves an existing model, evaluated sequentially, while the
stepwise search asks whether the complete model, considered as a
whole, justifies its total complexity. Linkage groups 4, 6, and 10
were supported by both approaches and are considered the most
reliable QTL identified in this study. Linkage groups 5 and 8 are
reported as genuine but more marginal findings, each individually
significant within the five-QTL model, but more sensitive to the
specific statistical criterion applied. Future work with a larger
population or denser marker coverage in these regions may help
resolve whether these two loci represent stable, repeatable QTL or
estimates closer to the threshold of detection.

No evidence of epistatic interaction was found among any of the
identified QTL, whether tested specifically among the five loci in
the final model or across the genome as a whole. Within the
limitations of the population size and marker density available
here, this suggests that plant height in the Z020 population is
controlled by a set of additive loci acting largely independently
of one another, rather than through interactive or epistatic gene
networks. This has a practical implication for breeding: favourable
alleles at these loci, particularly the NC350 alleles at linkage
groups 4, 6, 8, and 10, could in principle be combined through
marker-assisted selection without needing to account for
unfavourable allele combinations at interacting loci, simplifying
the design of a selection strategy aimed at increasing or
decreasing plant height as a breeding objective.

The absence of significant QTL on six of the ten maize chromosomes
(linkage groups 1, 2, 3, 7, and 9, after accounting for linkage
group 5's modest signal) is consistent with a moderate-effect,
oligogenic architecture for plant height rather than a highly
polygenic trait controlled by very many loci of small effect
scattered across the genome. This pattern aligns with previous
findings in the maize nested association mapping panel, where
plant height and related architecture traits have repeatedly been
shown to be controlled by a relatively small number of QTL of
moderate to large effect, alongside a larger number of
undetectable, very small effect loci that individual mapping
populations of this size are not well powered to detect.

Several limitations of this study are worth noting. The linkage map
was constructed from a single recombinant inbred line population of
188 individuals, which, while adequate for detecting the moderate
to large effect QTL reported here, has limited power to detect
loci of small effect or to resolve closely linked QTL into separate
peaks. Residual heterozygosity, present in approximately 5.5% of
genotype calls and treated as missing data throughout this
analysis, may have introduced a small amount of additional noise
into recombination fraction and QTL effect estimates, though this
is standard practice for recombinant inbred lines at this stage of
inbreeding and is unlikely to have materially altered the major
findings reported here. Finally, this analysis considered plant
height in isolation; given that linkage group 6 and other loci
identified here may also influence correlated traits such as ear
height or leaf architecture, a multi-trait analysis would be a
valuable extension of this work.

---

# References

*To be managed via references.bib. Key citations to add: McMullen et al.
2009 (NAM panel), Taniguti et al. 2023 (OneMap), Lander et al. 1987
(MAPMAKER), Broman et al. 2003 (R/qtl), Gaynor et al. 2021 (AlphaSimR).*

---

# Linkage Mapping

## Load Required Packages

```{r load-packages}
# onemap does the linkage map construction: segregation testing,
# two-point rf, grouping, ordering, HMM distances, map plots
#
# if not installed, run once in console (not in a knitted chunk):
#   install.packages("devtools")
#   devtools::install_github("Cristianetaniguti/onemap",
#                             dependencies = TRUE, upgrade = TRUE)

library(onemap)
```

---

## Load and Inspect the Raw Data

```{r load-data}
# Z020 (NC350 x B73) RIL data, csv lives in data/
#
# 188 rows = individuals (Z020E001-Z020E188)
# 1144 cols total:
#   col 1          = Ind, used as row names
#   cols 2-1107    = 1106 SNP markers, 0 = NC350 (AA), 1 = het, 2 = B73 (BB)
#   cols 1108-1144 = 38 phenotype columns
# ~2624 missing genotype values, no missing phenotypes

raw_data <- read.csv(
  file      = "data/Z020 NC350 x B73.csv",
  header    = TRUE,
  row.names = 1      # Ind becomes row names
)

cat("Rows (individuals):", nrow(raw_data), "\n")
cat("Columns (markers + phenotypes):", ncol(raw_data), "\n")
```

---

## Separate Markers from Phenotypes

```{r separate-markers-phenotypes}
# split the csv into geno and pheno: a column counts as a marker if
# every non-missing value is 0, 1, or 2
#
# PlantHeight (col 13 of phenotypes) is the trait I'm mapping

is_marker_col <- function(x) {
  non_missing <- x[!is.na(x)]
  all(non_missing %in% c(0, 1, 2))
}

marker_flags <- sapply(raw_data, is_marker_col)
marker_cols  <- names(raw_data)[marker_flags]
pheno_cols   <- names(raw_data)[!marker_flags]

cat("Marker columns identified:", length(marker_cols), "\n")
cat("Phenotype columns identified:", length(pheno_cols), "\n")
cat("\nPhenotype column names:\n")
print(pheno_cols)

geno_mat     <- as.matrix(raw_data[, marker_cols])
phenotype_df <- raw_data[, pheno_cols]

M <- ncol(geno_mat)
N <- nrow(geno_mat)

plant_height        <- phenotype_df[["PlantHeight"]]
names(plant_height) <- rownames(raw_data)

cat("\nPlantHeight summary:\n")
print(summary(plant_height))
```

---

## Descriptive Statistics for Plant Height

```{r plant-height-distribution, fig.cap="Figure 1. Distribution of plant height (cm) across 188 Z020 NC350 x B73 RIL individuals. The red curve is the fitted normal distribution."}
# plant height = base to tassel tip, quantitative trait, expect
# roughly normal spread across the RILs
#
# normality matters here because the regression-based QTL methods
# later assume normal residuals, hence the Shapiro-Wilk check below

hist(
  plant_height,
  breaks = 20,
  main   = "Distribution of Plant Height in Z020 RIL Population",
  xlab   = "Plant Height (cm)",
  ylab   = "Frequency",
  col    = "steelblue",
  border = "white"
)

curve(
  dnorm(x, mean = mean(plant_height), sd = sd(plant_height)) *
    length(plant_height) * diff(range(plant_height)) / 20,
  add = TRUE,
  col = "firebrick",
  lwd = 2
)

legend(
  "topright",
  legend = c("Observed", "Normal curve"),
  fill   = c("steelblue", NA),
  lty    = c(NA, 1),
  col    = c("steelblue", "firebrick"),
  lwd    = c(NA, 2),
  border = c("black", NA),
  bty    = "n"
)

sw        <- shapiro.test(plant_height)
cat("\nShapiro-Wilk normality test for PlantHeight:\n")
cat("  W =", round(sw$statistic, 4), "\n")
cat("  p-value =", round(sw$p.value, 4), "\n")
if (sw$p.value > 0.05) {
  cat("  No significant departure from normality (p > 0.05).\n")
} else {
  cat("  Significant departure from normality detected (p < 0.05).",
      "Consider a log or square-root transformation before QTL mapping.\n")
}
```

---

## Convert Genotype Matrix to MAPMAKER Format

```{r convert-to-raw}
# onemap reads MAPMAKER/EXP .raw format, so converting the numeric
# csv matrix to that first
#
# cross type "ri self": RILs from repeated selfing of a single F2,
# ~F6-F7 here, ~5.5% residual het, no heterozygote class in the model
#
# recoding: 0 (NC350) -> A, 2 (B73) -> B, 1 (het) -> missing,
# NA -> missing. Het calls go to missing because ri self has no het
# class and 5.5% is little enough to lose without real cost

cat("Individuals (N):", N, "\n")
cat("Markers (M):", M, "\n")

n_het <- sum(geno_mat == 1, na.rm = TRUE)
cat("Residual heterozygous calls recoded as missing:", n_het,
    "(", round(100 * n_het / (N * M), 2), "% of all cells)\n")

onemap_mat <- matrix("", nrow = N, ncol = M,
                     dimnames = dimnames(geno_mat))
onemap_mat[geno_mat == 0]   <- "A"
onemap_mat[geno_mat == 2]   <- "B"
onemap_mat[geno_mat == 1]   <- "-"   # het -> missing
onemap_mat[is.na(geno_mat)] <- "-"   # missing -> missing

marker_names     <- paste0("*", colnames(onemap_mat))
genotype_strings <- apply(onemap_mat, 2, paste, collapse = " ")
marker_lines     <- paste(marker_names, genotype_strings, sep = " ")

header <- c(
  "data type ri self",
  paste(N, M, 0)
)

writeLines(c(header, marker_lines), "data/Z020_NC350_x_B73.raw")

cat("MAPMAKER .raw file written: data/Z020_NC350_x_B73.raw\n")
cat("Lines written:", 2 + M, "(2 header + 1 per marker)\n")
```

---

## Load Data into OneMap

```{r load-into-onemap, fig.cap="Figure 2. Genotype dot plot for all 188 RIL individuals across all markers. Light dots = AA (NC350), dark dots = BB (B73), black dots = missing. A salt-and-pepper pattern confirms unbiased data loading."}
# read_mapmaker() reads the .raw file into a onemap object, printing
# it confirms cross type/individuals/markers came in correctly
#
# the plot should look like random salt-and-pepper, no stripes -
# stripes would mean a block of individuals missing data across a
# whole region, a data quality flag

my_data <- read_mapmaker(file = "data/Z020_NC350_x_B73.raw")

print(my_data)

plot(my_data)

save(my_data, file = "data/my_data.RData")
cat("\nmy_data saved to data/my_data.RData\n")
```

---

## Marker Segregation Testing

```{r segregation-test, fig.cap="Figure 3. Chi-square segregation test results for all markers. Markers to the right of the dashed orange Bonferroni threshold line show significant deviation from the expected 1:1 AA:BB ratio."}
# expected 1:1 AA:BB at every marker for a self-derived RIL
#
# test_segregation() runs chi-square per marker, Bonferroni_alpha()
# adjusts the threshold for testing all M markers at once
# (~0.000045 for M = 1106)
#
# distorted markers aren't automatically a problem here - some
# distortion is normal in RILs from divergent parents (selection
# during selfing, genotyping artefacts, real fitness differences)

bc_test          <- test_segregation(my_data)
bonferroni_alpha <- Bonferroni_alpha(bc_test)

cat("Bonferroni-corrected significance threshold:", bonferroni_alpha, "\n")

plot(bc_test)

n_distorted <- sum(bc_test$Pvalue < bonferroni_alpha, na.rm = TRUE)
cat("Significantly distorted markers:", n_distorted,
    "(", round(100 * n_distorted / M, 1), "%)\n")
```

---

## Two-Point Recombination Fractions

```{r twopoint-rf}
# rf_2pts() = pairwise recombination fraction for every marker pair,
# 1106 markers -> 610,765 pairs, takes 5-20 min, don't close R
#
# LOD >= 3 between a pair means linked is 1000x more likely than
# unlinked (r = 0.5)
#
# saving immediately once done - if R crashes before save this whole
# calc has to run again

rdata_twopts <- "data/bc_twopts.RData"

if (file.exists(rdata_twopts)) {
  load(rdata_twopts)
  cat("Loaded two-point rf results from:", rdata_twopts, "\n")
} else {
  cat("Computing two-point rf for", M * (M - 1) / 2,
      "marker pairs. Please wait...\n")
  bc_twopts <- rf_2pts(my_data)
  save(bc_twopts, file = rdata_twopts)
  cat("Done. Results saved to:", rdata_twopts, "\n")
}

cat("\nSample output (markers 1 and 2):\n")
print(bc_twopts, c(1, 2))
```

---

## Suggest LOD Threshold for Grouping

```{r lod-suggestion}
# suggest_lod() recommends a grouping threshold based on N and M -
# a data-driven cutoff beats the fixed LOD = 3 rule of thumb, since
# a denser/larger population like this can detect weaker linkage
# reliably, so the threshold needs to be higher to avoid merging
# unlinked markers by mistake

LOD_sug <- suggest_lod(my_data)
cat("Suggested LOD threshold for grouping:", round(LOD_sug, 3), "\n")
```

---

## Marker Grouping into Linkage Groups

```{r marker-grouping, fig.cap="Figure 4. UPGMA dendrogram showing hierarchical clustering of all markers. Each coloured cluster enclosed in a red rectangle is one linkage group, corresponding to one of the 10 maize chromosomes."}
# grouping markers on two criteria together: LOD >= LOD_sug and
# rf <= 0.35
#
# maize = 10 chromosome pairs, so expecting 10 groups out of this.
# if it comes out wrong: lower LOD merges groups, higher LOD splits
# a chromosome into pieces
#
# group_upgma() clusters on the rf distance matrix and gives the
# dendrogram - low rf merges happen near the bottom (tight linkage),
# unlinked markers only merge near rf = 0.5 at the top

bc_all_markers <- make_seq(bc_twopts, "all")

# direct LOD + max.rf assignment
bc_lgs <- group(bc_all_markers, max.rf = 0.35, LOD = LOD_sug)
cat("\n--- Grouping result (LOD + max.rf method) ---\n")
print(bc_lgs)

# UPGMA clustering, gives the dendrogram
bc_lgs_upgma <- group_upgma(
  bc_all_markers,
  expected.groups = 10,
  inter           = FALSE
)
cat("\n--- Grouping result (UPGMA method) ---\n")
print(bc_lgs_upgma)

plot(bc_lgs_upgma)
```

---

## Set the Kosambi Mapping Function

```{r mapping-function}
# converting recombination fraction (0-0.5) to genetic distance (cM)
#
# using Kosambi: d = 0.25 * ln[(1+2r)/(1-2r)], assumes positive
# crossover interference (one crossover suppresses nearby ones) -
# more realistic for maize than Haldane, gives shorter/tighter map
# length estimates
#
# this sets the mapping function for the whole session, has to run
# before any onemap::map() call below

set_map_fun(type = "kosambi")
cat("Mapping function set to: Kosambi\n")
```

---

## Extract Linkage Groups as Sequence Objects

```{r extract-lg-sequences}
# group() stores assignments in bc_lgs$groups - named integer vector,
# names = marker index, value = LG number (0 = unassigned)
#
# so max value (excluding 0/NA) = number of groups found
#
# make_seq(bc_lgs, i) pulls all markers in group i as a sequence
# object ready for ordering

cat("Class of bc_lgs:", class(bc_lgs), "\n")
cat("Names in bc_lgs object:", paste(names(bc_lgs), collapse = ", "), "\n\n")

group_vec <- bc_lgs$groups

cat("Type of bc_lgs$groups:", typeof(group_vec), "\n")
cat("Length:", length(group_vec), "\n\n")

n_assigned   <- sum(group_vec > 0, na.rm = TRUE)
n_unassigned <- sum(group_vec == 0, na.rm = TRUE)
cat("Markers assigned to a linkage group:", n_assigned, "\n")
cat("Markers unassigned (group = 0):", n_unassigned, "\n\n")

n_groups <- max(group_vec, na.rm = TRUE)
cat("Number of linkage groups found:", n_groups, "\n\n")

lg_seqs <- vector("list", n_groups)
for (i in seq_len(n_groups)) {
  lg_seqs[[i]] <- make_seq(bc_lgs, i)
}

cat("Number of markers per linkage group:\n")
for (i in seq_len(n_groups)) {
  cat("  LG", i, ":", length(lg_seqs[[i]]$seq.num), "markers\n")
}
```

---

## Filter Redundant Markers with find_bins

```{r filter-redundant-markers}
# markers with identical genotype patterns across all 188
# individuals add nothing to the map. order_seq() chokes on lots of
# duplicate-pattern markers and spits out impossible chromosome
# lengths (saw LG7 ~483 cM, LG9 ~972 cM before filtering)
#
# find_bins() run once on the full dataset, groups markers by
# identical/near-identical pattern, keep the first marker per bin as
# representative and drop the rest before ordering. dropped markers
# can be added back later at their bin position with add_marker()
# if needed
#
# exact = FALSE allows near-identical (up to 1 mismatch) patterns to
# bin together - sensible given we have missing calls

bins_all <- find_bins(my_data, exact = FALSE)

bin_representatives <- unlist(lapply(bins_all$bins, function(b) b[1]))

cat("Total bins (unique segregation patterns):",
    length(bins_all$bins), "\n")
cat("Representative markers retained:", length(bin_representatives), "\n")
cat("Redundant markers removed:",
    M - length(bin_representatives), "\n\n")

lg_seqs_filtered <- vector("list", n_groups)
for (i in seq_len(n_groups)) {
  lg_markers          <- lg_seqs[[i]]$seq.num
  filtered_markers    <- intersect(lg_markers, bin_representatives)
  lg_seqs_filtered[[i]] <- make_seq(bc_twopts, filtered_markers)
  cat("LG", i, ": reduced from", length(lg_markers),
      "to", length(filtered_markers), "markers after bin filtering\n")
}
```

---

## Order Markers and Estimate Map for All Linkage Groups

```{r order-all-lgs}
# order_seq() = branch and bound ordering per linkage group,
# make_seq(..., "safe") keeps only confidently-placed markers,
# onemap::map() then runs the HMM for multipoint rf, stored in seq.rf
#
# this onemap version doesn't give cumulative cM positions directly
# (no seq.pos), only per-interval rf in seq.rf, so converting to cM
# manually with Kosambi and taking the cumulative sum myself

rf_to_kosambi_cM <- function(rf) {
  0.25 * log((1 + 2 * rf) / (1 - 2 * rf)) * 100
}

bc_map      <- vector("list", n_groups)
bc_ord_safe <- vector("list", n_groups)

for (lg in seq_len(n_groups)) {

  cat("\n--- Linkage Group", lg, "---\n")
  rdata_file <- paste0("data/bc_lg", lg, "_ord.RData")

  if (file.exists(rdata_file)) {

    load(rdata_file)   # restores lg_ord_safe
    cat("  Loaded cached order from:", rdata_file, "\n")

  } else {

    cat("  Running order_seq() for LG", lg, "with",
        length(lg_seqs_filtered[[lg]]$seq.num), "markers...\n")
    cat("  This may take several minutes. Please wait.\n")

    lg_ord_raw <- order_seq(
      input.seq     = lg_seqs_filtered[[lg]],
      n.init        = 5,
      subset.search = "twopt",
      twopt.alg     = "rcd",
      THRES         = 3
    )

    lg_ord_safe <- make_seq(lg_ord_raw, "safe")
    save(lg_ord_safe, file = rdata_file)
    cat("  Order saved to:", rdata_file, "\n")
  }

  bc_ord_safe[[lg]] <- lg_ord_safe

  bc_map[[lg]] <- onemap::map(lg_ord_safe)

  rf_intervals <- bc_map[[lg]]$seq.rf
  cM_intervals <- rf_to_kosambi_cM(rf_intervals)
  total_cM     <- sum(cM_intervals, na.rm = TRUE)
  n_markers    <- length(bc_map[[lg]]$seq.num)

  cat("  Safe markers in map :", n_markers, "\n")
  cat("  Map length (cM)     :", round(total_cM, 2), "\n")
  cat("  Log-likelihood      :", round(bc_map[[lg]]$seq.like, 2), "\n")
}

save(bc_map, bc_ord_safe, file = "data/bc_map.RData")
cat("\n=== All linkage groups complete ===\n")
cat("Complete map saved to data/bc_map.RData\n")

cat("\n=== Map Length Summary ===\n")
cat(sprintf("%-6s  %-10s  %-12s  %-16s\n",
            "LG", "Markers", "Length(cM)", "AvgSpacing(cM)"))
cat(strrep("-", 50), "\n")

total_mrk <- 0
total_len <- 0

for (lg in seq_len(n_groups)) {
  n_mrk        <- length(bc_map[[lg]]$seq.num)
  cM_intervals <- rf_to_kosambi_cM(bc_map[[lg]]$seq.rf)
  len_cM       <- sum(cM_intervals, na.rm = TRUE)
  avg_sp       <- len_cM / (n_mrk - 1)

  cat(sprintf("%-6d  %-10d  %-12.1f  %-16.2f\n",
              lg, n_mrk, len_cM, avg_sp))

  total_mrk <- total_mrk + n_mrk
  total_len <- total_len + len_cM
}

cat(strrep("-", 50), "\n")
cat(sprintf("%-6s  %-10d  %-12.1f\n", "Total", total_mrk, total_len))
```

---

## Draw the Recombination Fraction Heatmap for LG1

```{r rf-heatmap-lg1, fig.cap="Figure 5. Recombination fraction heatmap for Linkage Group 1 after ordering. The clean red diagonal band confirms that markers placed close together on the map have low recombination fractions between them."}
# rf_graph_table() = colour-coded pairwise rf matrix for one LG
#
# correct ordering should give a clean dark red diagonal (adjacent
# markers tightly linked), off-diagonal yellow/green = markers far
# apart on the chromosome. a patchy/broken diagonal would flag
# ordering problems

rf_graph_table(bc_ord_safe[[1]], mrk.axis = "numbers")
```

---

## Draw the Genetic Linkage Map for LG1

```{r draw-map-lg1, fig.cap="Figure 6. Genetic linkage map for Linkage Group 1 of the Z020 NC350 x B73 RIL population. Each tick mark represents one SNP marker. Distances are in Kosambi centimorgans."}
# draw_map(): standard linkage map diagram - chromosome bar, tick
# per marker, y-axis = cM from the top marker
#
# names = TRUE prints marker labels, cex.mrk = 0.5 keeps labels small
# enough not to overlap on a dense chromosome

draw_map(bc_map[[1]], names = TRUE, grid = TRUE, cex.mrk = 0.5)
```

---

## Haplotype Reconstruction for LG1

```{r haplotype-lg1, fig.cap="Figure 7. Haplotype blocks for four representative RIL individuals across Linkage Group 1. Red = NC350 allele (AA); blue = B73 allele (BB). Colour transitions mark historical recombination breakpoints that occurred during the selfing process."}
# each RIL descends from one F2 plant via repeated selfing -
# heterozygous stretches get replaced by fixed NC350/B73 blocks over
# generations, colour transitions here mark where those crossovers
# happened during selfing
#
# expect a few large homozygous blocks per chromosome at F6-F7, with
# breakpoints scattered rather than clustered. where the two
# haplotype tracks disagree, that's one of the ~5.5% residual hets
#
# progeny_haplotypes() needs the ordered SEQUENCE object
# (bc_ord_safe), not the map object (bc_map) - map objects don't
# carry the raw genotypes needed for this

progeny_haplot <- progeny_haplotypes(
  bc_ord_safe[[1]],
  most_likely = TRUE,
  ind         = 1:4,
  group_names = "LG1"
)

plot(progeny_haplot, position = "split")
```

---

## Draw the Complete 10-Chromosome Linkage Map

```{r full-map, fig.cap="Figure 8. Complete genetic linkage map of the Z020 NC350 x B73 RIL population across all 10 maize chromosomes. Distances are in Kosambi centimorgans. Marker names are suppressed for clarity on dense chromosomes."}
# draw_map() takes the full bc_map list and draws all 10 LGs side by
# side - names = FALSE here since printing marker names on chromosomes
# with hundreds of markers each just turns into a black smear

draw_map(bc_map, names = FALSE, grid = TRUE, cex.mrk = 0.7)
```

---

## Map Summary Statistics

```{r map-summary}
# Table 1 source data - chromosome length from seq.rf via Kosambi,
# same approach as the order-all-lgs chunk since this onemap version
# doesn't store seq.pos directly

rf_to_kosambi_cM <- function(rf) {
  0.25 * log((1 + 2 * rf) / (1 - 2 * rf)) * 100
}

cat("=== Linkage Map Summary (Table 1) ===\n\n")
cat(sprintf("%-6s  %-10s  %-12s  %-16s\n",
            "LG", "Markers", "Length(cM)", "AvgSpacing(cM)"))
cat(strrep("-", 50), "\n")

total_markers <- 0
total_length  <- 0

for (lg in seq_len(n_groups)) {
  n_mrk        <- length(bc_map[[lg]]$seq.num)
  cM_intervals <- rf_to_kosambi_cM(bc_map[[lg]]$seq.rf)
  len_cM       <- sum(cM_intervals, na.rm = TRUE)
  avg_sp       <- len_cM / (n_mrk - 1)

  cat(sprintf("%-6d  %-10d  %-12.1f  %-16.2f\n",
              lg, n_mrk, len_cM, avg_sp))

  total_markers <- total_markers + n_mrk
  total_length  <- total_length  + len_cM
}

cat(strrep("-", 50), "\n")
cat(sprintf("%-6s  %-10d  %-12.1f\n",
            "Total", total_markers, total_length))
```

---

## Export Map for QTL Analysis

```{r export-map}
# write_map() exports the map in R/qtl-compatible format - LG
# assignments, marker order, Kosambi cM positions
#
# PlantHeight also saved separately as csv, both files feed into
# read.cross() for the QTL mapping sections below

write_map(bc_map, file.out = "data/bc.map")
cat("Linkage map exported to: data/bc.map\n")

ph_df <- data.frame(
  Ind         = names(plant_height),
  PlantHeight = as.numeric(plant_height),
  row.names   = NULL
)
write.csv(ph_df, "data/plant_height.csv", row.names = FALSE)
cat("Plant height data saved to: data/plant_height.csv\n")
cat("Both files are ready for QTL mapping.\n")
```

---

# Session Information

```{r session-info}
# R and package versions, for reproducibility
sessionInfo()
```

---

# QTL Mapping

## Load Required Packages

```{r qtl-load-packages}
# R/qtl handles everything from here on: SMA, interval mapping,
# composite interval mapping, multiple interval mapping
#
# install once in console if needed: install.packages("qtl")

library(qtl)
```

## Build the R/qtl Cross Object

```{r load-cross-object}
# read.cross(format = "mm") won't accept "data type ri self" in the
# header - it wants "data type backcross" or "data type f2
# intercross", with the real cross type passed separately via
# crosstype
#
# fix: write a copy of the .raw file with the header line swapped to
# "data type backcross" (R/qtl accepts this as a placeholder for mm
# format), then pass crosstype = "riself" to read.cross() so the
# correct genetic model still gets applied. genotype data itself is
# untouched, only the header line changes

raw_lines <- readLines("data/Z020_NC350_x_B73.raw")

cat("Original header line:\n")
cat(raw_lines[1], "\n\n")

raw_lines[1] <- "data type backcross"

writeLines(raw_lines, "data/Z020_qtl.raw")
cat("Corrected .raw file written to data/Z020_qtl.raw\n")
cat("New header line:", raw_lines[1], "\n\n")

z020 <- read.cross(
  format    = "mm",
  file      = "data/Z020_qtl.raw",
  mapfile   = "data/bc.map",
  crosstype = "riself"
)

z020$pheno$PlantHeight <- plant_height[
  match(rownames(z020$pheno), names(plant_height))
]

cat("\nCross object loaded successfully.\n")
print(summary(z020))
```

## Filter Genotype Data to Match the Final Map

```{r filter-geno-to-map}
# z020 above was still built off the unfiltered marker set - need a
# .raw file containing only the 242 markers that survived ordering,
# matching data/bc.map exactly

final_marker_names <- unlist(lapply(bc_map, function(x) {
  colnames(geno_mat)[x$seq.num]
}))

cat("Total markers in final ordered map:", length(final_marker_names), "\n")

geno_mat_filtered <- geno_mat[, final_marker_names]

N_filt <- nrow(geno_mat_filtered)
M_filt <- ncol(geno_mat_filtered)

onemap_mat_filt <- matrix("", nrow = N_filt, ncol = M_filt,
                          dimnames = dimnames(geno_mat_filtered))
onemap_mat_filt[geno_mat_filtered == 0]   <- "A"
onemap_mat_filt[geno_mat_filtered == 2]   <- "B"
onemap_mat_filt[geno_mat_filtered == 1]   <- "-"
onemap_mat_filt[is.na(geno_mat_filtered)] <- "-"

marker_names_filt     <- paste0("*", colnames(onemap_mat_filt))
genotype_strings_filt <- apply(onemap_mat_filt, 2, paste, collapse = " ")
marker_lines_filt     <- paste(marker_names_filt, genotype_strings_filt,
                               sep = " ")

header_filt <- c(
  "data type backcross",
  paste(N_filt, M_filt, 0)
)

writeLines(c(header_filt, marker_lines_filt), "data/Z020_qtl_filtered.raw")

cat("Filtered .raw file written: data/Z020_qtl_filtered.raw\n")
cat("Contains exactly", M_filt, "markers matching data/bc.map\n")
```

## Reload with the Filtered Marker Set

```{r reload-filtered-cross}
# rebuilding z020 from the filtered .raw file so it lines up with
# bc.map (242 markers), and restoring individual names by POSITION
# since pull.geno() / read.cross() don't carry IDs through - matching
# by name here would silently fail given how R/qtl handles this mm
# format, position-based restoration is the reliable way

load("data/my_data.RData")
load("data/bc_map.RData")

ph_df        <- read.csv("data/plant_height.csv")
plant_height <- ph_df$PlantHeight
names(plant_height) <- ph_df$Ind

raw_data <- read.csv(
  file      = "data/Z020 NC350 x B73.csv",
  header    = TRUE,
  row.names = 1
)

is_marker_col <- function(x) {
  non_missing <- x[!is.na(x)]
  all(non_missing %in% c(0, 1, 2))
}
marker_flags <- sapply(raw_data, is_marker_col)
marker_cols  <- names(raw_data)[marker_flags]
geno_mat     <- as.matrix(raw_data[, marker_cols])

library(onemap)
library(qtl)

z020 <- read.cross(
  format    = "mm",
  file      = "data/Z020_qtl_filtered.raw",
  mapfile   = "data/bc.map",
  crosstype = "riself"
)

# restore PlantHeight by position, not match() - raw_data and z020
# hold the same 188 individuals in the same original order
z020$pheno$PlantHeight <- plant_height[rownames(raw_data)]

cat("All objects reloaded successfully.\n")
cat("bc_map groups:", length(bc_map), "\n")
cat("z020 individuals:", nrow(z020$pheno), "\n")
```

## Single Marker Analysis

```{r genotype-phenotype-image, fig.cap="Figure 9. Genotype matrix sorted by plant height. Vertical banding indicates marker-trait association."}
# quick visual before the formal test: sort individuals by height,
# look for vertical banding in the genotype matrix lined up the same
# way
#
# no real association -> genotypes look like random salt-and-pepper
# regardless of sort order. real association -> a colour band
# clusters among the tallest or shortest plants at that marker
#
# pull.geno() doesn't carry row names through from the mm format
# reader, so restoring them by position from raw_data (same order,
# confirmed earlier)

geno_matrix <- pull.geno(z020)
rownames(geno_matrix) <- rownames(raw_data)

ph_sorted   <- sort(plant_height)
sorted_idx  <- match(names(ph_sorted), rownames(geno_matrix))

cat("Matched individuals:", sum(!is.na(sorted_idx)), "of", length(sorted_idx), "\n\n")

layout(matrix(c(1, 2, 2, 2), nrow = 1, ncol = 4, byrow = TRUE))

image(
  t(as.matrix(ph_sorted)),
  col  = rainbow(n = 256, end = 2/3, rev = TRUE),
  axes = FALSE,
  xlab = "Plant Height (cm)",
  ylab = paste("Individuals (N =", nrow(geno_matrix), ")")
)

image(
  t(geno_matrix[sorted_idx, ]),
  col  = c("yellow", "blue"),
  axes = FALSE,
  xlab = paste("Markers (M =", ncol(geno_matrix), ")")
)
```

```{r sma-linear-regression}
# fitting PlantHeight ~ marker genotype for every marker
# independently:
#
#   PlantHeight_i = mu + beta*x_i + epsilon_i
#
#   x_i coded 0 (NC350 homozygote) / 1 (B73 homozygote)
#   beta = additive effect of B73 allele vs NC350
#
# LOD converted from the F statistic:
#   LOD = (N/2) * log10(F * 1/(N-2) + 1)
# this puts SMA on the same LOD scale as IM/CIM/MIM later, so
# results are directly comparable across sections
#
# Bonferroni threshold uses alpha/M for the 242 markers in the final
# map
#
# rownames restored by position again, same reasoning as above -
# match() on this data silently drops everything to NA, learned that
# the hard way earlier in the project

geno_sma  <- pull.geno(z020)
rownames(geno_sma) <- rownames(raw_data)
pheno_sma <- z020$pheno$PlantHeight

geno_sma  <- geno_sma - 1   # AA -> 0, BB -> 1

n_markers_sma <- ncol(geno_sma)
N_sma         <- length(pheno_sma)

cat("Markers tested:", n_markers_sma, "\n")
cat("Individuals:", N_sma, "\n\n")

res_sma <- matrix(
  NA,
  nrow     = n_markers_sma,
  ncol     = 7,
  dimnames = list(
    colnames(geno_sma),
    c("Intercept", "Effect(cm)", "t value", "P value",
      "F value", "LOD score", "R-squared")
  )
)

for (m in seq_len(n_markers_sma)) {

  marker_geno <- geno_sma[, m]
  df_temp     <- data.frame(PH = pheno_sma, G = marker_geno)
  df_temp     <- df_temp[complete.cases(df_temp), ]

  if (nrow(df_temp) < 10) next

  fit <- summary(lm(PH ~ G, data = df_temp))
  n_i <- nrow(df_temp)

  LOD <- (n_i / 2) * log10(
    fit$fstatistic[1] * (1 / (n_i - 1 - 1)) + 1
  )

  res_sma[m, ] <- c(
    fit$coefficients[1, 1],
    fit$coefficients[2, 1],
    fit$coefficients[2, 3],
    fit$coefficients[2, 4],
    fit$fstatistic[1],
    LOD,
    fit$r.squared
  )
}

res_sma <- res_sma[complete.cases(res_sma), ]

bonf_p   <- 0.05 / nrow(res_sma)
bonf_lod <- min(
  res_sma[res_sma[, "P value"] <= bonf_p, "LOD score"],
  na.rm = TRUE
)

cat("Bonferroni P threshold:", signif(bonf_p, 3), "\n")
cat("Corresponding LOD threshold:", round(bonf_lod, 3), "\n\n")

sig_markers <- res_sma[res_sma[, "P value"] < bonf_p, , drop = FALSE]
cat("Significant markers after Bonferroni correction:",
    nrow(sig_markers), "\n\n")

if (nrow(sig_markers) > 0) {
  cat("Top significant markers (sorted by LOD score):\n")
  print(
    round(
      sig_markers[order(sig_markers[, "LOD score"],
                        decreasing = TRUE), ],
      4
    )
  )
}

save(res_sma, bonf_lod, sig_markers,
     file = "data/sma_results.RData")
cat("\nSMA results saved to data/sma_results.RData\n")
```

```{r sma-lod-plot, fig.cap="Figure 10. Single marker analysis LOD scores across all 10 linkage groups, in map order. The dashed red line marks the Bonferroni significance threshold."}
# plotting LOD by cumulative map position across all 10 LGs, with
# chromosome boundaries marked - lets the whole genome scan be read
# as one continuous figure rather than 10 separate ones

chr_map   <- pull.map(z020)
lod_df    <- data.frame(
  marker   = rownames(res_sma),
  LOD      = res_sma[, "LOD score"],
  stringsAsFactors = FALSE
)

lod_df$chr <- NA
lod_df$pos <- NA

for (ch in seq_along(chr_map)) {
  mrk_names <- names(chr_map[[ch]])
  idx       <- lod_df$marker %in% mrk_names
  lod_df$chr[idx] <- ch
  lod_df$pos[idx] <- chr_map[[ch]][
    match(lod_df$marker[idx], mrk_names)
  ]
}

chr_lengths  <- sapply(chr_map, max)
chr_offsets  <- c(0, cumsum(chr_lengths[-length(chr_lengths)]))
lod_df$cum_pos <- lod_df$pos + chr_offsets[lod_df$chr]

lod_df <- lod_df[order(lod_df$cum_pos), ]

plot(
  lod_df$cum_pos, lod_df$LOD,
  type  = "h",
  col   = "steelblue",
  lwd   = 1.2,
  xlab  = "Map position (cM)",
  ylab  = "LOD score",
  main  = "Single Marker Analysis -- Plant Height",
  xaxt  = "n",
  ylim  = c(0, max(lod_df$LOD, bonf_lod * 1.2, na.rm = TRUE))
)

for (ch in seq_along(chr_map)) {
  mid_pos <- chr_offsets[ch] + chr_lengths[ch] / 2
  axis(1, at = mid_pos, labels = paste0("LG", ch), las = 2, cex.axis = 0.7)
  if (ch < length(chr_map)) {
    abline(v = chr_offsets[ch] + chr_lengths[ch],
           col = "grey70", lty = 2, lwd = 0.8)
  }
}

abline(h = bonf_lod, col = "firebrick", lty = 2, lwd = 1.5)
text(
  x      = max(lod_df$cum_pos, na.rm = TRUE) * 0.02,
  y      = bonf_lod + max(lod_df$LOD, na.rm = TRUE) * 0.03,
  labels = paste0("Threshold (LOD = ", round(bonf_lod, 2), ")"),
  col    = "firebrick",
  cex    = 0.75,
  adj    = 0
)
```

```{r sma-scanone-mr}
# cross-checking the manual lm() loop above against R/qtl's own
# marker regression (scanone method = "mr") - if these two
# independent routes agree, that's good confidence the manual loop
# and the LOD conversion formula are both correct

z020.mr <- scanone(
  z020,
  pheno.col = "PlantHeight",
  method    = "mr"
)

cat("scanone() marker regression complete.\n\n")
cat("Markers with LOD above Bonferroni threshold (",
    round(bonf_lod, 3), "):\n")
print(summary(z020.mr, threshold = bonf_lod))

save(z020.mr, file = "data/z020_mr.RData")

plot(
  z020.mr,
  ylab = "LOD score",
  main = "Single Marker Analysis (scanone) -- Plant Height",
  type = "p",
  col  = "steelblue"
)
abline(h = bonf_lod, col = "firebrick", lty = 2, lwd = 1.5)
```

```{r sma-top-marker-plot}
# pulling out the top SMA marker for a closer look, plotPXG() shows
# the phenotype-by-genotype distribution directly

top_marker <- rownames(z020.mr)[which.max(z020.mr$lod)]

cat("Top marker from SMA:", top_marker, "\n")
cat("LOD score:", round(max(z020.mr$lod, na.rm = TRUE), 3), "\n")
cat("Chromosome:", z020.mr$chr[which.max(z020.mr$lod)], "\n")
cat("Position (cM):", round(z020.mr$pos[which.max(z020.mr$lod)], 2), "\n\n")

if (top_marker %in% rownames(res_sma)) {
  cat("Additive effect (BB - AA):",
      round(res_sma[top_marker, "Effect(cm)"], 3), "cm\n")
  cat("R-squared:",
      round(res_sma[top_marker, "R-squared"], 4), "\n")
  cat("Variance explained:",
      round(res_sma[top_marker, "R-squared"] * 100, 2), "%\n")
}

plotPXG(z020, top_marker, pheno.col = "PlantHeight")
```

```{r sma-fitqtl}
# fitqtl() at the top SMA marker, using multiple imputation
# (n.draws = 32) for a proper effect size estimate rather than just
# trusting the nearest-marker lm() coefficient
#
# step = 0 here since this is a marker-position-only fit, not a
# between-marker interval estimate

z020 <- sim.geno(z020, n.draws = 32, step = 0)

top_chr <- z020.mr$chr[which.max(z020.mr$lod)]
top_pos <- z020.mr$pos[which.max(z020.mr$lod)]

cat("Fitting QTL at:\n")
cat("  Marker    :", top_marker, "\n")
cat("  Chromosome:", top_chr, "\n")
cat("  Position  :", round(top_pos, 2), "cM\n\n")

z020.qtl <- makeqtl(z020, chr = top_chr, pos = top_pos)

z020.fit <- fitqtl(
  z020,
  qtl     = z020.qtl,
  formula = y ~ Q1,
  get.ests = TRUE,
  pheno.col = "PlantHeight"
)

print(summary(z020.fit))

save(z020.qtl, z020.fit, top_marker, top_chr, top_pos,
     file = "data/sma_fitqtl.RData")
cat("\nfitqtl results saved to data/sma_fitqtl.RData\n")
```

## Single Marker Analysis Results Summary

Single marker analysis tested all 242 markers in the final linkage
map for association with plant height, identifying 13 markers
significant after Bonferroni correction. The strongest signal, on
linkage group 6, explained nearly 16% of phenotypic variance on its
own, a substantial proportion for a single marker and an early
indication that this region carries a major QTL for plant height.
Secondary signals on linkage groups 4 and 5 were weaker but
consistent with the pattern later confirmed by more sophisticated
mapping methods. The one unexpected result was the antagonistic
marker on linkage group 10, where the B73 allele rather than NC350
was associated with greater height, a finding that warranted closer
attention in later sections. The absence of any significant markers
on six of the ten chromosomes pointed towards a moderate-effect
genetic architecture rather than a highly polygenic one, a pattern
that held up throughout the rest of the analysis.

---

## Interval Mapping

```{r calc-genoprob}
# IM needs genotype probabilities at every position along the map,
# not just at the markers themselves - calc.genoprob() does this
# using the flanking markers and the Kosambi distances already in
# z020
#
# step = 1 -> a probability calculated every 1 cM, on top of the
# marker positions, ~1550 testing positions across the ~1549 cM map
#
# riself means only 2 QTL genotype classes possible at any position
# (NC350/NC350 or B73/B73) - simpler than the backcross/F2 cases with
# 2-4 classes

z020 <- calc.genoprob(z020, step = 1)

cat("Genotype probabilities calculated.\n")
cat("Step size: 1 cM\n")
cat("Number of chromosomes:", nchr(z020), "\n")
cat("Total map length (cM):", round(sum(sapply(pull.map(z020), max)), 1), "\n")
```

```{r scanone-im}
# scanone() at every genoprob position, fitting plant height against
# QTL genotype prob and recording LOD - running both:
#
#   em : exact ML via Expectation-Maximization, slower, the gold
#        standard hk is approximating
#   hk : Haley-Knott regression, fast least-squares approximation,
#        holds up well with moderate missing data and decent marker
#        spacing
#
# if em and hk profiles match closely, that's good evidence the
# result isn't an artefact of the estimation method chosen

z020.em <- scanone(z020, pheno.col = "PlantHeight", method = "em")
cat("EM scan complete.\n")

z020.hk <- scanone(z020, pheno.col = "PlantHeight", method = "hk")
cat("HK scan complete.\n\n")

cat("--- Top position, EM method ---\n")
print(z020.em[which.max(z020.em$lod), ])

cat("\n--- Top position, HK method ---\n")
print(z020.hk[which.max(z020.hk$lod), ])

save(z020.em, z020.hk, file = "data/im_scanone.RData")
cat("\nSaved to data/im_scanone.RData\n")
```

```{r im-permutation-test}
# permutation test for the genome-wide significance threshold -
# shuffle PlantHeight across individuals 1000 times, breaking the
# real genotype-phenotype link, keep the single highest LOD from each
# shuffled scan. 95th percentile of those 1000 max LODs = alpha 0.05
# threshold
#
# running both methods to match what gets compared later:
#   hk : fast, runs directly
#   em : slow on a 1549 cM map x 1000 perms, cached to disk so a
#        session restart doesn't mean starting over
#
# set.seed() so both runs are reproducible if re-run

set.seed(678)
z020.perm.hk <- scanone(
  z020,
  pheno.col = "PlantHeight",
  method    = "hk",
  n.perm    = 1000,
  verbose   = FALSE
)
cat("HK permutations complete.\n")

rdata_perm_em <- "data/im_perm_em.RData"

if (file.exists(rdata_perm_em)) {

  load(rdata_perm_em)
  cat("Loaded cached EM permutation results from:", rdata_perm_em, "\n")

} else {

  cat("Running 1000 EM permutations. This will take a while.\n")
  cat("R will save the result automatically when done.\n")

  set.seed(678)
  z020.perm.em <- scanone(
    z020,
    pheno.col = "PlantHeight",
    method    = "em",
    n.perm    = 1000,
    verbose   = FALSE
  )

  save(z020.perm.em, file = rdata_perm_em)
  cat("EM permutations complete. Saved to:", rdata_perm_em, "\n")
}

cat("\n--- LOD threshold at alpha = 0.05 ---\n")
cat("HK method:\n")
print(summary(z020.perm.hk, alpha = 0.05))

cat("EM method:\n")
print(summary(z020.perm.em, alpha = 0.05))

save(z020.perm.hk, z020.perm.em,
     file = "data/im_perm_both.RData")
cat("\nBoth permutation objects saved to data/im_perm_both.RData\n")
```

```{r im-results-summary}
# summary() with perms= returns only positions clearing the
# permutation threshold at the chosen alpha - one row per peak
#
# note: summary() on a scanone perm object returns a named numeric
# vector, so pulling the threshold needs ["lod"], not $lod

thres_em <- summary(z020.perm.em, alpha = 0.05)["lod"]
thres_hk <- summary(z020.perm.hk, alpha = 0.05)["lod"]

cat("=== Significant QTL positions from Interval Mapping ===\n\n")

cat("--- EM method (threshold LOD =", round(thres_em, 2), ") ---\n")
print(summary(z020.em, perms = z020.perm.em, alpha = 0.05))

cat("\n--- HK method (threshold LOD =", round(thres_hk, 2), ") ---\n")
print(summary(z020.hk, perms = z020.perm.hk, alpha = 0.05))
```

```{r im-lod-plot, fig.cap="Figure 11. Interval mapping LOD profiles across the genome, comparing the EM (red) and HK (blue) methods. Dashed lines mark each method's genome-wide significance threshold."}
# overlaying em and hk LOD profiles plus their thresholds - should
# sit almost on top of each other, any visible gap is where missing
# genotype data makes the hk approximation drift slightly from the
# exact em solution

plot(
  z020.em, z020.hk,
  ylab  = "LOD score",
  main  = "Interval Mapping -- Plant Height",
  col   = c("red", "blue"),
  lwd   = 2
)

add.threshold(z020.em, perms = z020.perm.em, col = "red",   lty = 2)
add.threshold(z020.hk, perms = z020.perm.hk, col = "blue",  lty = 2)

legend(
  "topright",
  legend  = c("EM", "HK"),
  lty     = 1,
  lwd     = 2,
  col     = c("red", "blue"),
  box.col = NA
)
```

```{r im-fitqtl}
# fitqtl() at the IM peak (LG6, 108 cM), using multiple imputation
# so the effect estimate is attributed to the inferred QTL position
# rather than whichever marker happens to be nearest
#
# re-running sim.geno() with step = 1 here to match the IM grid - the
# earlier sim.geno() call used step = 0 for SMA and that draw set
# doesn't apply to interval positions
#
# riself -> fitqtl() fits 2 classes (NC350/NC350 vs B73/B73), and the
# additive effect it reports is HALF the homozygote difference -
# different convention to the lm() effect from SMA, which is the
# FULL difference. multiply fitqtl's estimate by 2 to compare the two

z020 <- sim.geno(z020, n.draws = 32, step = 1)

im_top_chr <- 6
im_top_pos <- 108

z020.im.qtl <- makeqtl(z020, chr = im_top_chr, pos = im_top_pos)

cat("QTL object built at:\n")
cat("  Chromosome:", im_top_chr, "\n")
cat("  Position  :", im_top_pos, "cM\n\n")
print(z020.im.qtl)

z020.im.fit <- fitqtl(
  z020,
  qtl       = z020.im.qtl,
  formula   = y ~ Q1,
  get.ests  = TRUE,
  pheno.col = "PlantHeight"
)

print(summary(z020.im.fit))

save(z020.im.qtl, z020.im.fit,
     file = "data/im_fitqtl.RData")
cat("\nfitqtl results saved to data/im_fitqtl.RData\n")
```

## Interval Mapping Results Summary

Interval mapping confirmed the linkage group 6 signal as the
dominant QTL for plant height, with its LOD score rising from 6.62
under single marker analysis to 7.34, reflecting the added
resolution gained from testing positions between markers rather
than only at marker locations. Three further regions on linkage
groups 4, 5, and 10 cleared the genome-wide significance threshold,
matching the candidate regions flagged by single marker analysis.
The close agreement between the EM and HK estimation methods across
the whole genome scan gave confidence that these results were not
sensitive to the choice of algorithm. Taken together with the single
marker analysis, this stage of the project confirmed four candidate
QTL regions worth carrying forward into composite interval mapping,
where background genetic variation could be properly controlled for.

---

## Reload Point: Restoring Objects for a New Session

```{r reload-saved-objects}
# run this at the start of any new R session to bring back every
# object needed for the rest of the report, without rerunning any of
# the slow computation (two-point rf, marker ordering, permutations)

library(onemap)
library(qtl)

raw_data <- read.csv(
  file      = "data/Z020 NC350 x B73.csv",
  header    = TRUE,
  row.names = 1
)

is_marker_col <- function(x) {
  non_missing <- x[!is.na(x)]
  all(non_missing %in% c(0, 1, 2))
}
marker_flags <- sapply(raw_data, is_marker_col)
marker_cols  <- names(raw_data)[marker_flags]
geno_mat     <- as.matrix(raw_data[, marker_cols])

ph_df        <- read.csv("data/plant_height.csv")
plant_height <- ph_df$PlantHeight
names(plant_height) <- ph_df$Ind

load("data/my_data.RData")    # my_data
load("data/bc_twopts.RData")  # bc_twopts
load("data/bc_map.RData")     # bc_map, bc_ord_safe

z020 <- read.cross(
  format    = "mm",
  file      = "data/Z020_qtl_filtered.raw",
  mapfile   = "data/bc.map",
  crosstype = "riself"
)
z020$pheno$PlantHeight <- plant_height[rownames(raw_data)]

z020 <- calc.genoprob(z020, step = 1)
z020 <- sim.geno(z020, n.draws = 32, step = 1)

load("data/sma_results.RData")   # res_sma, bonf_lod, sig_markers
load("data/z020_mr.RData")       # z020.mr
load("data/sma_fitqtl.RData")    # z020.qtl, z020.fit, top_marker, top_chr, top_pos

load("data/im_scanone.RData")    # z020.em, z020.hk
load("data/im_perm_both.RData")  # z020.perm.em, z020.perm.hk
load("data/im_fitqtl.RData")     # z020.im.qtl, z020.im.fit

cat("All objects restored successfully.\n")
cat("bc_map groups:", length(bc_map), "\n")
cat("z020 individuals:", nrow(z020$pheno), "\n")
cat("Total markers in map:", sum(sapply(pull.map(z020), length)), "\n")
```

---

## Composite Interval Mapping

```{r cim-load-packages-check}
# cim() lives in qtl, already loaded - just confirming the objects
# needed from interval mapping are in the environment before going
# further. if starting fresh, run reload-saved-objects above first

library(qtl)

stopifnot(exists("z020"))
stopifnot(exists("z020.em"))
stopifnot(exists("z020.perm.em"))

cat("z020 cross object present:", exists("z020"), "\n")
cat("Individuals:", nind(z020), "\n")
cat("Total markers:", totmar(z020), "\n")
cat("Chromosomes:", nchr(z020), "\n")
cat("Genotype probabilities calculated:",
    "prob" %in% names(z020$geno[[1]]), "\n")

cat("\nReady to proceed with Composite Interval Mapping.\n")
```

```{r cim-window15}
# cim() tests plant height against the QTL genotype probability at
# each position, plus a set of marker cofactors elsewhere in the
# genome that soak up the effect of other QTL so they don't blur the
# test at the current position
#
# n.marcovar = 27, following 2*sqrt(N) for N = 188 - matches the
# professor's value exactly since our population is the same size
#
# window = 15: cofactors within 15 cM of the tested position get
# excluded, so a cofactor can't accidentally absorb the very QTL
# being tested
#
# method = "hk" - cim() is heavier than plain scanone, hk keeps it
# practical
#
# running this straight on z020, no convert2riself() needed since
# ours was built as riself from the start (the professor's tutorial
# converts because his file was read in as a backcross)

set.seed(678)
z020.cim15 <- cim(
  cross      = z020,
  pheno.col  = "PlantHeight",
  method     = "hk",
  n.marcovar = 27,
  window     = 15
)

cat("CIM with window = 15 cM complete.\n\n")

cat("Peak LOD per chromosome:\n")
print(summary(z020.cim15))

save(z020.cim15, file = "data/cim_window15.RData")
cat("\nSaved to data/cim_window15.RData\n")
```

```{r cim-window-inf}
# same cim() run but window = Inf - the strictest setting, no
# cofactor allowed anywhere on the SAME linkage group as the tested
# position, only cofactors from other chromosomes
#
# useful as a check against the window = 15 result: a peak holding up
# under window = Inf is good evidence it's real and not an artefact
# of a nearby cofactor distorting the local model. a peak that
# shrinks a lot under window = Inf needs more caution

set.seed(678)
z020.cimInf <- cim(
  cross      = z020,
  pheno.col  = "PlantHeight",
  method     = "hk",
  n.marcovar = 27,
  window     = Inf
)

cat("CIM with window = Inf complete.\n\n")

cat("Peak LOD per chromosome:\n")
print(summary(z020.cimInf))

save(z020.cimInf, file = "data/cim_windowInf.RData")
cat("\nSaved to data/cim_windowInf.RData\n")
```

```{r cim-permutation}
# same permutation logic as IM, just on cim() now - shuffle
# PlantHeight 1000 times, record the max LOD per shuffle, 95th
# percentile = alpha 0.05 threshold
#
# running the permutation at window = Inf only: it's the most
# conservative setting so the resulting threshold is valid across
# both window sizes, and running just one permutation set instead of
# two saves a lot of time given this is the slowest step in the CIM
# section
#
# cached to disk, this one's slow enough that a session restart
# shouldn't mean starting over

rdata_perm_cim <- "data/cim_perm.RData"

if (file.exists(rdata_perm_cim)) {

  load(rdata_perm_cim)
  cat("Loaded cached CIM permutation results from:", rdata_perm_cim, "\n")

} else {

  cat("Running 1000 CIM permutations with window = Inf.\n")
  cat("This is the slowest step in the CIM section. Please wait.\n")

  set.seed(678)
  z020.perm.cim <- cim(
    cross      = z020,
    pheno.col  = "PlantHeight",
    method     = "hk",
    n.marcovar = 27,
    window     = Inf,
    n.perm     = 1000
  )

  save(z020.perm.cim, file = rdata_perm_cim)
  cat("CIM permutations complete. Saved to:", rdata_perm_cim, "\n")
}

cim_threshold <- summary(z020.perm.cim, alpha = 0.05)
cat("\n--- CIM LOD threshold at alpha = 0.05 ---\n")
print(cim_threshold)
```

```{r cim-apply-threshold}
# applying the LOD 6.24 threshold to both window scans to see which
# peaks are genuinely significant rather than just locally tall
#
# summary() with perms= only returns positions clearing the
# threshold, max one row per chromosome (the highest peak that
# clears it)

cat("=== Significant QTL from CIM, window = 15 ===\n")
cim15_sig <- summary(z020.cim15, perms = z020.perm.cim, alpha = 0.05)
print(cim15_sig)

cat("\n=== Significant QTL from CIM, window = Inf ===\n")
cimInf_sig <- summary(z020.cimInf, perms = z020.perm.cim, alpha = 0.05)
print(cimInf_sig)

save(cim15_sig, cimInf_sig, cim_threshold,
     file = "data/cim_significant.RData")
cat("\nSaved to data/cim_significant.RData\n")
```

```{r cim-plot, fig.cap="Figure 12. Composite interval mapping LOD profiles across all 10 linkage groups, comparing window = 15 cM (orange) and window = infinite (cyan) cofactor exclusion settings. The dashed red line marks the genome-wide significance threshold (LOD 6.24, alpha = 0.05). Green ticks mark positions of marker cofactors selected for the window = 15 scan."}
# both CIM window scans overlaid, threshold drawn as a horizontal
# line, green ticks via add.cim.covar() showing where the window = 15
# cofactors were picked - comparing window settings visually makes it
# obvious which peaks are stable vs sensitive to the window choice

plot(
  z020.cim15, z020.cimInf,
  col  = c("orange", "cyan"),
  main = "Composite Interval Mapping -- Plant Height"
)

add.threshold(
  z020.cim15,
  perms = z020.perm.cim,
  alpha = 0.05,
  col   = "firebrick",
  lty   = 2
)

add.cim.covar(z020.cim15, col = "darkgreen")

legend(
  "topright",
  legend = c("window = 15", "window = Inf", "Threshold (LOD 6.24)"),
  lty    = c(1, 1, 2),
  col    = c("orange", "cyan", "firebrick"),
  box.col = NA
)
```

```{r cim-makeqtl}
# building the QTL object from the 3 CIM-confirmed positions (LG4,
# LG6, LG8), using the window = Inf positions since that's the
# stricter scan and its peaks are a touch more stable (no
# same-chromosome cofactor can nudge them)
#
# sim.geno() re-run here so the imputation draws match the current
# genoprob grid - same n.draws = 32, step = 1 as the IM fitqtl step

z020 <- sim.geno(z020, n.draws = 32, step = 1)

cim_chr <- cimInf_sig$chr
cim_pos <- cimInf_sig$pos

cat("Building QTL object at:\n")
for (i in seq_along(cim_chr)) {
  cat("  Q", i, ": LG", cim_chr[i], " at ", round(cim_pos[i], 1),
      " cM (LOD ", round(cimInf_sig$lod[i], 2), ")\n", sep = "")
}

z020.cim.qtl <- makeqtl(z020, chr = cim_chr, pos = cim_pos)
print(z020.cim.qtl)

save(z020.cim.qtl, cim_chr, cim_pos, file = "data/cim_qtl_object.RData")
cat("\nSaved to data/cim_qtl_object.RData\n")
```

```{r cim-fitqtl-combined}
# fitting all 3 QTL together rather than one at a time:
#
#   - joint %var properly accounts for any shared variance between
#     QTL instead of double-counting overlap from 3 separate
#     single-QTL fits
#   - the drop-one-at-a-time table shows what each QTL contributes
#     ABOVE the other two, a fairer measure of independent
#     importance than a stand-alone effect
#
# formula = y ~ Q1 + Q2 + Q3, additive only, no interaction term -
# epistasis testing is MIM's job later, not CIM's

z020.cim.fit <- fitqtl(
  z020,
  qtl       = z020.cim.qtl,
  formula   = y ~ Q1 + Q2 + Q3,
  get.ests  = TRUE,
  pheno.col = "PlantHeight"
)

cat("=== Combined three-QTL model (CIM) ===\n\n")
print(summary(z020.cim.fit))

save(z020.cim.fit, file = "data/cim_fitqtl.RData")
cat("\nSaved to data/cim_fitqtl.RData\n")
```

```{r cim-vs-im-plot, fig.cap="Figure 13. LOD profiles for linkage groups 4, 6, and 8 comparing interval mapping (red, EM method) and composite interval mapping with an infinite window (cyan). Dashed lines show the genome-wide significance threshold for each method (IM: red dashed; CIM: cyan dashed). Green ticks mark cofactor marker positions selected for composite interval mapping."}
# IM vs CIM, restricted to the 3 chromosomes CIM confirmed (LG4, 6,
# 8) - cutting the rest of the genome out of the plot makes the
# sharpening effect of CIM much easier to actually see
#
# using z020.em for IM (exact ML, the method treated as primary IM
# result) and z020.cimInf for CIM (stricter window, the one the QTL
# model above is built from)

plot(
  z020.em, z020.cimInf,
  chr  = c(4, 6, 8),
  col  = c("red", "cyan"),
  main = "Interval Mapping versus Composite Interval Mapping"
)

add.threshold(
  z020.em,
  perms = z020.perm.em,
  alpha = 0.05,
  col   = "red",
  lty   = 2
)

add.threshold(
  z020.cimInf,
  perms = z020.perm.cim,
  alpha = 0.05,
  col   = "cyan",
  lty   = 2
)

add.cim.covar(z020.cimInf, chr = c(4, 6, 8), col = "darkgreen")

legend(
  "topleft",
  legend  = c("IM (EM method)", "CIM (window = Inf)",
              "IM threshold", "CIM threshold"),
  lty     = c(1, 1, 2, 2),
  col     = c("red", "cyan", "red", "cyan"),
  box.col = NA
)
```

## Composite Interval Mapping Results Summary

Composite interval mapping confirmed three QTL for plant height,
all consistent across two cofactor exclusion window settings
(15 cM and infinite). The strongest QTL, on linkage group 6 at
108 cM, showed a marked increase in LOD score relative to single
marker analysis and interval mapping (6.62, 7.34, and 17.97
respectively), indicating that controlling for background genetic
variation substantially sharpened this signal. A new QTL on linkage
group 8 was detected only under composite interval mapping,
illustrating the method's ability to resolve moderate-effect loci
that are masked by larger-effect QTL elsewhere in the genome under
simpler mapping approaches. Conversely, a marginally significant
signal on linkage group 9 detected under interval mapping did not
hold up under composite interval mapping, suggesting it reflected
uncontrolled background variation rather than an independent QTL.
Jointly, the three confirmed QTL (linkage groups 4, 6, and 8)
explained 33.87% of phenotypic variance in plant height. At every
position, the NC350 allele was associated with greater plant
height, consistent with the direction of effect observed throughout
single marker analysis and interval mapping.

---

## Multiple Interval Mapping

```{r mim-load-and-check}
# MIM picks up directly from CIM - z020 already has calc.genoprob()
# and sim.geno() attached, no new cross object or map needed
#
# just confirming the CIM objects are still around before going
# further (skipping the professor's summaryMap/plotMap/plotRF checks
# here since the map was already verified back in the linkage mapping
# section, no need to repeat that)

library(qtl)

stopifnot(exists("z020"))
stopifnot(exists("z020.cim.qtl"))   # the 3-QTL object from CIM (LG4, 6, 8)
stopifnot(exists("z020.cim.fit"))   # the fitqtl() result from CIM

cat("Starting Multiple Interval Mapping.\n")
cat("Carrying forward CIM-confirmed QTL: LG4, LG6, LG8\n\n")

print(z020.cim.qtl)
```

```{r mim-refineqtl}
# refineqtl() re-optimises each QTL position one at a time while
# holding the others fixed - a QTL's best position can shift slightly
# once the others are properly accounted for jointly, which CIM's
# cofactor approach doesn't quite do (cofactors control for
# background variance but don't jointly optimise QTL positions
# against each other)
#
# iterates through Q1/Q2/Q3 testing small shifts, repeats until no
# more LOD gain, prints its own progress as it goes

z020.ref <- refineqtl(
  cross     = z020,
  pheno.col = "PlantHeight",
  qtl       = z020.cim.qtl,
  formula   = y ~ Q1 + Q2 + Q3
)

cat("\nRefined QTL positions:\n")
print(z020.ref)

save(z020.ref, file = "data/mim_ref1.RData")
cat("\nSaved to data/mim_ref1.RData\n")
```

```{r mim-scantwo-permutations}
# scantwo() tests every PAIR of genome positions jointly, not single
# positions like scanone/cim - needed here for two things:
#
#   1. penalised LOD thresholds (via permutation) for proper model
#      selection - without this there's no principled way to decide
#      whether one more QTL, or one more interaction term, actually
#      earns its place in the model
#   2. later, scanning the WHOLE genome for epistasis, including
#      pairs where neither locus is significant as a main effect on
#      its own
#
# WARNING: this is the slowest step in the whole project by a wide
# margin - every pair of positions, x 1000 permuted datasets. can run
# for hours. cached to disk like the other slow steps. do NOT
# interrupt this once it starts or it has to begin again from scratch

rdata_perm_two <- "data/mim_scantwo_perm.RData"

if (file.exists(rdata_perm_two)) {

  load(rdata_perm_two)
  cat("Loaded cached scantwo permutation results from:",
      rdata_perm_two, "\n")

} else {

  cat("Running 1000 scantwo permutations. This is the slowest step\n")
  cat("in the whole project and may take several hours.\n")
  cat("Do not close R or restart your session while this runs.\n\n")

  set.seed(678)
  z020.perm.two <- scantwo(
    z020,
    pheno.col = "PlantHeight",
    method    = "hk",
    n.perm    = 1000
  )

  save(z020.perm.two, file = rdata_perm_two)
  cat("\nscantwo permutations complete. Saved to:", rdata_perm_two, "\n")
}

cat("\n--- scantwo permutation thresholds (alpha = 0.05) ---\n")
print(summary(z020.perm.two, alpha = 0.05))
```

```{r mim-fitqtl-refined}
# re-fitting the 3-QTL model at the REFINED positions (LG4 @ 67,
# LG6 @ 108, LG8 @ 126) rather than the original CIM peaks - since
# positions moved slightly under refineqtl(), the test stats need
# recalculating at the new positions

z020.fit.ref <- fitqtl(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.ref,
  formula   = y ~ Q1 + Q2 + Q3,
  get.ests  = TRUE
)

cat("=== Refined three-QTL model ===\n\n")
print(summary(z020.fit.ref))

save(z020.fit.ref, file = "data/mim_fitqtl_ref1.RData")
cat("\nSaved to data/mim_fitqtl_ref1.RData\n")
```

```{r mim-addqtl-round1}
# addqtl() scans the genome asking: if a QTL got added HERE, on top
# of the 3 already in the model, how much would the model improve?
# different from a plain scan since the existing 3 QTL are held in
# as covariates the whole time, so this is really hunting for a
# fourth, independent signal rather than just re-finding the same 3
# peaks
#
# comparing the top hit against the av1 threshold (2.55) from the
# scantwo permutations - that's the threshold calibrated specifically
# for "is one more additive QTL worth adding to an existing model"

z020.add1 <- addqtl(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.ref,
  formula   = y ~ Q1 + Q2 + Q3
)

cat("=== addqtl scan: searching for a 4th QTL ===\n\n")
cat("Top candidate position per chromosome:\n")
print(summary(z020.add1))

cat("\nOverall maximum:\n")
print(max(z020.add1))

cat("\nav1 threshold (from scantwo permutations):", 2.55, "\n")

top_add1_lod <- max(z020.add1)$lod
if (top_add1_lod > 2.55) {
  cat("Result: LOD", round(top_add1_lod, 2),
      "exceeds the threshold. This position is a candidate to add.\n")
} else {
  cat("Result: LOD", round(top_add1_lod, 2),
      "does not exceed the threshold. No further QTL to add.\n")
}

save(z020.add1, file = "data/mim_addqtl1.RData")
cat("\nSaved to data/mim_addqtl1.RData\n")
```

```{r mim-add-fourth-qtl}
# adding the LG10 candidate (102 cM) to the 3-QTL model and
# refining again - a new QTL can shift where the others sit best,
# so always refine right after adding
#
# NOTE: max(z020.add1)$chr comes back as an R factor. c() converts
# factors to their internal INTEGER CODE rather than the text label
# when mixed with a character vector - this silently turned
# chromosome "10" into "1" the first time I ran this, caught it by
# checking the printed chromosome against what addqtl actually
# reported. as.character() wrapped around the factor fixes it, kept
# here as a standing reminder for the rest of the addqtl rounds below

chr <- c(as.character(max(z020.add1)$chr), z020.ref$chr)
pos <- c(max(z020.add1)$pos, z020.ref$pos)

cat("Building 4-QTL object at:\n")
for (i in seq_along(chr)) {
  cat("  Q", i, ": LG", chr[i], " at ", round(pos[i], 1), " cM\n", sep = "")
}

z020.qtl2 <- makeqtl(z020, chr = chr, pos = pos)
print(z020.qtl2)

z020.ref2 <- refineqtl(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.qtl2,
  formula   = y ~ Q1 + Q2 + Q3 + Q4
)

cat("\nRefined 4-QTL positions:\n")
print(z020.ref2)

save(z020.qtl2, z020.ref2, file = "data/mim_qtl2_ref2.RData")
cat("\nSaved to data/mim_qtl2_ref2.RData\n")
```

```{r mim-fitqtl-4qtl}
# fitting the 4-QTL model (LG4, LG6, LG8, LG10) - this is also where
# the LG10 sign gets checked against the antagonistic marker
# PZA02386.2 flagged back in SMA, where B73 was associated with
# taller plants, opposite to every other significant marker at the
# time

z020.fit2 <- fitqtl(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.ref2,
  formula   = y ~ Q1 + Q2 + Q3 + Q4,
  get.ests  = TRUE
)

cat("=== Four-QTL model (LG4, LG6, LG8, LG10) ===\n\n")
print(summary(z020.fit2))

save(z020.fit2, file = "data/mim_fitqtl2.RData")
cat("\nSaved to data/mim_fitqtl2.RData\n")
```

```{r mim-addqtl-round2}
# second addqtl() round, now conditional on all 4 confirmed QTL -
# checking for a 5th independent signal, same logic as round 1

z020.add2 <- addqtl(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.ref2,
  formula   = y ~ Q1 + Q2 + Q3 + Q4
)

cat("=== addqtl scan: searching for a 5th QTL ===\n\n")
print(summary(z020.add2))

cat("\nOverall maximum:\n")
print(max(z020.add2))

cat("\nav1 threshold (from scantwo permutations):", 2.55, "\n")

top_add2_lod <- max(z020.add2)$lod
if (top_add2_lod > 2.55) {
  cat("Result: LOD", round(top_add2_lod, 2),
      "exceeds the threshold. This position is a candidate to add.\n")
} else {
  cat("Result: LOD", round(top_add2_lod, 2),
      "does not exceed the threshold. Stop here, no further QTL.\n")
}

save(z020.add2, file = "data/mim_addqtl2.RData")
cat("\nSaved to data/mim_addqtl2.RData\n")
```

```{r mim-add-fifth-qtl}
# adding the LG5 candidate (13 cM, LOD 2.65) - worth flagging this is
# a marginal addition, only just over the 2.55 threshold, not as
# clear-cut as LG4/LG6/LG8/LG10 were
#
# as.character() applied to the chromosome value again, same factor
# coercion issue as before, not taking any chances with it a second
# time

chr <- c(as.character(max(z020.add2)$chr), z020.ref2$chr)
pos <- c(max(z020.add2)$pos, z020.ref2$pos)

cat("Building 5-QTL object at:\n")
for (i in seq_along(chr)) {
  cat("  Q", i, ": LG", chr[i], " at ", round(pos[i], 1), " cM\n", sep = "")
}

z020.qtl3 <- makeqtl(z020, chr = chr, pos = pos)
print(z020.qtl3)

z020.ref3 <- refineqtl(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.qtl3,
  formula   = y ~ Q1 + Q2 + Q3 + Q4 + Q5
)

cat("\nRefined 5-QTL positions:\n")
print(z020.ref3)

save(z020.qtl3, z020.ref3, file = "data/mim_qtl3_ref3.RData")
cat("\nSaved to data/mim_qtl3_ref3.RData\n")
```

```{r mim-fitqtl-5qtl}
# fitting the 5-QTL model (LG4, LG5, LG6, LG8, LG10) - checking
# whether LG5 holds up as genuinely significant once properly fitted
# jointly, given it came in as a marginal addqtl() hit, and checking
# its effect size/direction relative to the other four

z020.fit3 <- fitqtl(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.ref3,
  formula   = y ~ Q1 + Q2 + Q3 + Q4 + Q5,
  get.ests  = TRUE
)

cat("=== Five-QTL model (LG4, LG5, LG6, LG8, LG10) ===\n\n")
print(summary(z020.fit3))

save(z020.fit3, file = "data/mim_fitqtl3.RData")
cat("\nSaved to data/mim_fitqtl3.RData\n")
```

```{r mim-addqtl-round3}
# third addqtl() round, conditional on all 5 confirmed QTL - checking
# whether a 6th turns up, or whether this is where the manual search
# naturally stops

z020.add3 <- addqtl(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.ref3,
  formula   = y ~ Q1 + Q2 + Q3 + Q4 + Q5
)

cat("=== addqtl scan: searching for a 6th QTL ===\n\n")
print(summary(z020.add3))

cat("\nOverall maximum:\n")
print(max(z020.add3))

cat("\nav1 threshold (from scantwo permutations):", 2.55, "\n")

top_add3_lod <- max(z020.add3)$lod
if (top_add3_lod > 2.55) {
  cat("Result: LOD", round(top_add3_lod, 2),
      "exceeds the threshold. This position is a candidate to add.\n")
} else {
  cat("Result: LOD", round(top_add3_lod, 2),
      "does not exceed the threshold. STOP. The manual search ends",
      "at five QTL.\n")
}

save(z020.add3, file = "data/mim_addqtl3.RData")
cat("\nSaved to data/mim_addqtl3.RData\n")
```

```{r mim-calc-penalties}
# stepwiseqtl() needs the penalty values in a specific format, not
# the summary table printed earlier - calc.penalties() converts the
# scantwo permutation results into the three values it actually uses:
#
#   main  : penalty per extra QTL (main effect)
#   heavy : interaction penalty, from the "full" threshold, used
#           comparing a 2-QTL interacting model against no QTL
#   light : interaction penalty, from the "fv1" threshold, used
#           comparing a 2-QTL interacting model against the best
#           single-QTL model
#
# these let stepwiseqtl() weigh whether a QTL or an interaction
# genuinely earns its complexity, same logic as the av1 threshold
# used by hand throughout the addqtl() rounds above

penalties <- calc.penalties(z020.perm.two)

cat("Penalties for stepwiseqtl():\n")
print(penalties)

save(penalties, file = "data/mim_penalties.RData")
cat("\nSaved to data/mim_penalties.RData\n")
```

```{r mim-stepwiseqtl}
# stepwiseqtl() runs an automatic forward/backward search: starts
# from no QTL, adds the best position, keeps adding (refining
# positions each time) as long as pLOD keeps improving, then works
# backward checking whether dropping any QTL still leaves an equal or
# better pLOD
#
# this is the independent cross-check against the manual search - if
# they agree, good confirmation the manual process found the right
# answer. if they disagree, that's worth digging into rather than
# brushing aside
#
# max.qtl = 8 gives some headroom above the manual 5-QTL result in
# case the automatic search lands somewhere different

z020.step <- stepwiseqtl(
  z020,
  pheno.col = "PlantHeight",
  max.qtl   = 8,
  method    = "hk",
  penalties = penalties
)

cat("\n=== stepwiseqtl() final model ===\n\n")
print(z020.step)

save(z020.step, file = "data/mim_stepwiseqtl.RData")
cat("\nSaved to data/mim_stepwiseqtl.RData\n")
```

```{r mim-addint}
# addint() tests every pairwise interaction among the QTL already in
# the model (LG4, LG5, LG6, LG8, LG10), asking whether adding that
# specific interaction term improves the model beyond the two main
# effects alone
#
# narrower test than the genome-wide scantwo() scan below - this only
# checks pairs among QTL already known, can't find an interacting
# partner with no main effect of its own
#
# relevant threshold here is "int" (4.01) from the scantwo
# permutations, since this is testing the interaction term
# specifically

z020.addint <- addint(
  z020,
  pheno.col = "PlantHeight",
  qtl       = z020.ref3,
  formula   = y ~ Q1 + Q2 + Q3 + Q4 + Q5
)

cat("=== Pairwise interaction tests among the 5 confirmed QTL ===\n\n")
print(z020.addint)

cat("\nint threshold (from scantwo permutations): 4.01\n")
max_int_lod <- max(z020.addint$LOD)
cat("Highest interaction LOD found:", round(max_int_lod, 3), "\n")

if (max_int_lod > 4.01) {
  cat("Result: at least one interaction exceeds the threshold.\n")
} else {
  cat("Result: no interaction among the 5 confirmed QTL exceeds the",
      "threshold. No evidence of epistasis among these specific loci.\n")
}

save(z020.addint, file = "data/mim_addint.RData")
cat("\nSaved to data/mim_addint.RData\n")
```

```{r mim-scantwo-actual}
# the scantwo permutation chunk above only computed the permuted null
# distribution - the scan on the REAL (non-permuted) data is a
# separate run, needed to actually identify interacting pairs once
# compared against the permutation thresholds
#
# cached to disk like the other slow steps, though this should be
# noticeably quicker than the 1000-permutation run since it's one
# scan, not a thousand

rdata_scantwo_actual <- "data/mim_scantwo_actual.RData"

if (file.exists(rdata_scantwo_actual)) {

  load(rdata_scantwo_actual)
  cat("Loaded cached scantwo actual-data results from:",
      rdata_scantwo_actual, "\n")

} else {

  cat("Running scantwo() on the actual data (single scan, not",
      "permuted). This should take a few minutes.\n")

  z020.two <- scantwo(
    z020,
    pheno.col = "PlantHeight",
    method    = "hk"
  )

  save(z020.two, file = rdata_scantwo_actual)
  cat("Saved to:", rdata_scantwo_actual, "\n")
}

cat("\n=== Genome-wide two-QTL scan summary ===\n")
cat("(showing pairs with notable full-model or interaction LOD)\n\n")
print(summary(z020.two))

save(z020.two, file = "data/mim_scantwo_actual.RData")
```

```{r mim-plot-lod-profile, fig.cap="Figure 14. LOD profile for the final five-QTL multiple interval mapping model, showing the profile around each QTL position with all other QTL held fixed in the model. Peak height and sharpness reflect each QTL's relative contribution and positional resolution."}
# plotLodProfile() shows the LOD profile around each QTL's position
# with the other four held fixed in the model - different from a
# plain genome scan since it shows how sharply each position is
# resolved GIVEN the rest of the model is already accounted for,
# which is the more relevant view once a multi-QTL model exists
#
# a narrow sharp peak = well-resolved position. a broad flat one
# usually points to lower marker density or a smaller effect size in
# that region

plotLodProfile(
  z020.ref3,
  main = "LOD Profile for Final 5-QTL Model -- Plant Height"
)
```

## Multiple Interval Mapping Results Summary

Multiple interval mapping built on the three QTL confirmed by
composite interval mapping, identifying two additional loci through
iterative scanning: linkage group 10 at 102 cM and linkage group 5
at 13 cM. The final model of five QTL explained 42.31% of phenotypic
variance in plant height, with the linkage group 6 and linkage
group 4 loci contributing the largest individual effects. An
independent automatic search confirmed linkage groups 4, 6, and 10
as the most robustly supported QTL, while linkage groups 5 and 8
were retained by the manual search as additional, more marginal
loci. The linkage group 5 QTL was notable for showing the opposite
direction of effect to all other loci in the model, with the B73
allele rather than the NC350 allele associated with greater plant
height. No evidence of epistatic interaction was found between any
pair of QTL, either among the five loci in the final model or
genome-wide, indicating that plant height in this population is
controlled by a small number of additive QTL of varying effect
size, consistent with the moderate-effect polygenic architecture
suggested by the trait's continuous distribution.

